Learning-Associated Flexibility of Cortical Taste Coding Is Impaired in Shank3 Knockout Mice

Version 1

Abstract

The ability to update the valence of sensory perception to influence behavior is crucial for survival. A common phenotype in autism spectrum disorders (ASDs) is defects in sensory processing, but whether these defects impair flexible sensory encoding is largely unexplored. In particular, how genetic risk factors such as Shank3 deletion affect the adaptability of cortical taste processing and downstream behavior is unknown. To address this gap, we performed two-photon calcium imaging during a conditioned taste aversion (CTA) learning paradigm, an ethologically relevant form of associative learning that depends on taste processing in the anterior insular cortex (AIC), to examine how Shank3 knockout alters taste-related neuronal activity in AIC and influences CTA learning. We found that AIC neurons in Shank3 knockout mice exhibited reduced stimulus-evoked suppression and increased trial-to-trial correlated variability during the acquisition of CTA memory. These activity changes, which likely reduced signal-to-noise ratio in AIC, were associated with slower CTA acquisition in knockout mice. In both genotypes, CTA learning enhanced, while subsequent extinction reduced, taste discriminability in AIC, and both extinction and the associated reduction in discriminability were faster in knockout than in wild-type mice. Together, these results show that Shank3 loss is associated with destabilized cortical activity dynamics in AIC, which may contribute to inefficient encoding and maintenance of learned taste aversion. These findings show that loss of Shank3 compromises the ability of animals to update behavior to incorporate negative outcomes, and suggest this loss of flexibility may be an important feature of monogenic ASDs.

Introduction

Ethological experiences can modify fundamental features of sensory coding, such as the valence of a taste, through changes in synaptic and network functions within sensory cortices (Harris and Mrsic-Flogel, 2013; Choi, 2018). This flexibility in sensory coding ensures that animals’ perceptions and downstream behaviors are updated to reflect current experiences and enhance survival. Sensory processing, including taste, is often defective in patients with autism spectrum disorders (ASDs) (Leekam et al., 2007; Avery et al., 2018; Scheerer et al., 2021; Goldschlager et al., 2025), and several rodent monogenic ASD models, including Shank3 knockout, exhibit sensory deficits resembling those observed in patients with ASD (Kloth et al., 2015; Copping et al., 2017; Monteiro and Feng, 2017; Delling and Boeckers, 2021). This altered sensory processing could contribute to downstream defects in flexible sensory coding in Shank3 knockout rodent models, but this possibility has not been directly tested. Here, we used conditioned taste aversion (CTA), an ethologically relevant form of associative learning that involves taste processing, to examine how disrupted cortical taste coding impacts the learning and extinction of aversive taste associations in Shank3 knockout mice.

CTA arises when a normally palatable taste is paired with agents (such as lithium chloride) that cause gastric malaise (Garcia et al., 1955); this aversive learning is fast (requiring one or two conditioning trials) and strongly suppresses the consumption of the conditioned tastant. Like other forms of aversive learning, CTA undergoes extinction when the tastant no longer predicts the adverse outcome (Berman and Dudai, 2001). Extinction occurs gradually over repeated exposures to the conditioned tastant alone; thus, during CTA and subsequent extinction, the valence of a tastant can shift from positive to negative and back to positive again in an ethologically relevant manner, providing a nice paradigm for studying the adaptability of sensory coding. CTA relies on the anterior insular cortex (AIC), where much of taste coding occurs (Katz et al., 2002; Fletcher et al., 2017; Wang et al., 2018; Levitan et al., 2019; Vincis et al., 2020; Chen et al., 2021). CTA’s acquisition and extinction induce distinct forms of plasticity within AIC (Escobar et al., 1998; Li et al., 2016; Rodríguez-Durán et al., 2017; Haley et al., 2020; Wu et al., 2021; Yiannakas et al., 2021; Kolatt Chandran et al., 2023) as internal representations of taste are reshaped (Moran and Katz, 2014; Lavi et al., 2018; Arieli et al., 2022; Raymond et al., 2024). Loss of Shank3 affects multisensory integration in the insular cortex (Gogolla et al., 2014) and impairs several cellular forms of cortical synaptic and intrinsic plasticity (Bozdagi et al., 2010; Wang et al., 2011; Kouser et al., 2013; Raynaud et al., 2013; Tatavarty et al., 2020; Wen and Turrigiano, 2021), raising the possibility that loss of Shank3 might impair CTA learning and extinction.

In this study, we employed two-photon calcium imaging and a head-restrained CTA paradigm to compare AIC activity in Shank3 knockout and wild-type mice during learning and extinction. We first discovered that Shank3 knockout mice learned CTA more slowly. During memory encoding, AIC neurons in these mutant mice were less suppressed by taste stimuli, with increased correlated variability that likely reduced the signal-to-noise ratio critical for taste processing within AIC. Although Shank3 knockout mice eventually formed CTA, their aversive memory was fragile and extinguished more rapidly than in wild-type mice. Following learning, AIC neurons in both groups became more reliable in their taste responsiveness; however, this reliability was accompanied by a gradual reduction in single-cell taste selectivity and a convergence of population activity as extinction progressed. This loss of taste selectivity and convergence of population activity was faster and more pronounced in Shank3 knockout than in wild-type mice. Finally, the ability to discriminate between tastants using population activity was enhanced by CTA and then maintained during extinction in wild-type mice but was rapidly lost during extinction in the Shank3 knockout condition. This work demonstrates that loss of Shank3 degrades the discriminability of single-cell and population activity in AIC, leading to slower learning and faster extinction of CTA. More broadly, this work shows that flexible sensory coding is defective in Shank3 knockout mice, and provides insight into how maladaptive sensory coding impacts associative learning in ASD.

Results

Shank3 knockout diminishes taste-elicited insular suppression and perturbs CTA acquisition

Loss of Shank3 is known to affect some aspects of sensory processing (Chen et al., 2020; Tatavarty et al., 2020; Balasco et al., 2022; Ortiz-Cruz et al., 2022), but whether it disrupts ethologically relevant associative learning paradigms such as CTA that rely on sensory processing is unknown. To examine this, we performed longitudinal two-photon calcium imaging of AIC neurons from wild-type (WT) and Shank3-knockout mice (KO, Δexon 13-16, Peça et al., 2011) in a head-restrained CTA learning paradigm (Figure 1A-C). Using a gradient-index (GRIN) lens (Miura et al., 2020; Deng et al., 2021; Staszko et al., 2022) and a virally expressed calcium indicator (jGCaMP8s) driven by the pan-neuronal promoter Synapsin (Zhang et al., 2023), we were able to visualize calcium signals of individual neurons within deep layers of the dysgranular and agranular AIC, subregions crucial for the expression of CTA (Figure 1B, C, Yiannakas et al., 2021; Kolatt Chandran et al., 2023).

Shank3 knockout diminishes taste-elicited insular suppression and perturbs CTA acquisition

(A) The head-restrained aversive taste learning paradigm. CS: conditioned stimulus, 25 mM saccharin, US: unconditioned stimulus, lithium chloride (LiCl, 0.15M, 2% body weight). CST: conditioning session, T: retrieval test session. (B) Example image showing the GRIN lens placement on top of the anterior insular cortex (AIC) (green: GCaMP8s, magenta: NeuN). Scale bar: 500 μm. (C) Example image showing the calcium signals recorded from individual AIC neurons. Scale bar: 50 μm. (D, E, F) Trial-averaged licks during the first (CST1) and second (CST2) conditioning sessions in the CS-only (WT) group (D), the CTA (WT) group (E), and the CTA (KO) group (F) (CS-only (WT): paired t-test, p = 0.1201; CTA (WT): paired t-test, p = 0.0094; CTA (KO): paired t-test, p = 0.0490). Here and below where behavior data are presented: CS-only (WT), n = 7 animals; CTA (WT), n = 9 animals; CTA (KO), n = 9 animals. (G) The trial-averaged lick ratio of CST2 to CST1 in CTA (WT) and CTA (KO) groups. (WT vs. KO: unpaired two-tailed t-test, p = 0.0123). (H, I) Heatmaps showing z-scored changes in calcium responses (ΔF/F) from neurons that are responsive to tastant delivery during conditioning in CTA (WT) (H, CST1: 189 cells, CST2: 172 cells) and CTA (KO) (I, CST1: 66 cells, CST2: 52 cells) groups. 1: cue onset; 2: spout in place; 3: tastant delivery starts; 4: tastant delivery ends. The 3-second taste delivery is marked by yellow dashed lines. In each heatmap, cells are sorted in descending temporal order based on whether their activity is activated or suppressed. Here and below where calcium imaging data are presented: CS-only (WT), n = 7 animals; CTA (WT), n = 8 animals; CTA (KO), n = 7 animals. (J) Calcium traces averaged across suppressed neurons in the CTA (WT) and CTA (KO) groups during CST1 (left) and CST2 (right). (K) Calcium traces averaged across activated neurons in the CTA (WT) and CTA (KO) groups during CST1 (left) and CST2 (right). (L) Change in the percentage of suppressed neurons from CST1 to CST2 (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.0218, session: p = 0.0393, interaction: 0.7819). (M) Change in the calcium fluorescence (ΔF/F) of suppressed neurons from CST1 to CST2. Dashed lines mark the value 0 (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.0011, session: p = 0.7356, interaction: 0.8672). (N) Change in the percentage of activated neurons from CST1 to CST2 (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.1391, session: p = 0.2590, interaction: 0.9705). (O) Change in the calcium fluorescence (ΔF/F) of activated neurons from CST1 to CST2 (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.3404, session: p = 0.2664, interaction: 0.1295). Error bars in this figure and below indicate s.e. (standard error of the mean).

We then trained water-restricted, head-restrained mice (WT or KO littermates) to lick a motorized spout in daily sessions, each consisting of approximately 30 trials (Chen et al., 2021). This trial-based structure enabled us to control the type of the tastant animals received in each trial, as well as the duration and timing of its delivery, while simultaneously recording the corresponding neuronal calcium signals in AIC (Figure 1A, C). To ensure that animals learned CTA effectively, we subjected habituated animals to two conditioning sessions (CST1 and CST2); within each session animals received, on average, 30 deliveries of the novel tastant saccharin (conditioned stimulus, CS), followed by a LiCl injection (unconditioned stimulus, US) (referred to as the “CTA group”). A separate group of animals received saline instead of LiCl and served as the “non-learned” comparison (referred to as the “CS-only group”). In subsequent retrieval test sessions that spanned five days (T1-5), we measured the expression and extinction of CTA by giving animals saccharin (the conditioned tastant) and water (a neutral tastant) in an alternative-choice task (Chen et al., 2021; Kogan and Fontanini, 2024). While animals in the CS-only group maintained their preference for saccharin across the first and second conditioning sessions (CST1 and CST2, Figure 1D), WT mice that received LiCl significantly reduced licks to saccharin between CST1 and CST2 (Figure 1E). Although KO animals also showed a significant reduction in their saccharin consumption between CST1 and CST2 (Figure 1F), this reduction was smaller than in WT mice (Figure 1G).

To ensure reduced CTA during CST2 in KO mice was not an artifact of measuring aversion without a choice between saccharin and water, we used a modified paradigm where a separate group of animals received an alternative-choice test after CST1 (Figure 1 – figure supplement 1A). In this paradigm as well, KO mice showed a significantly higher preference for saccharin across concentrations compared to WT littermates (Figure 1 – figure supplement 1B); further, comparable lick counts in WT and KO mice suggest no major differences in the thirst state or motivation between groups (Figure 1 – figure supplement 1C, D).

Since taste responses are processed in AIC (Katz et al., 2002), we wished to know whether saccharin-elicited activity in AIC during CTA acquisition is different in WT and KO animals. We identified neurons that were activated or suppressed during various phases of CST1 and 2, and (as observed previously, Chen et al., 2021) found that AIC neurons could be influenced by acoustic cues (1-2), lick movements (2-3), and tastant exposure (3-4, Figure 1 – figure supplement 2A-C). Most of these neurons responded to more than one sensorimotor stimulus (Figure 1 – figure supplement 2D), highlighting the multi-modal and integrative nature of the insular cortex (Samuelsen et al., 2012; Gogolla et al., 2014; Vincis and Fontanini, 2016; Gogolla, 2017). In most AIC neurons, the strongest trial-averaged responses were tied to tastant delivery (Figure 1H, I, Figure 1 – figure supplement 2A, B; between yellow dashed lines marked 3 and 4); in this study, we focus on these taste-elicited responses. In both WT and KO mice that underwent CTA, we detected neuronal populations that were either suppressed (Figure 1H, I, J) or activated (Figure 1H, I, K) by tastant presentations during CST1 and CST2. There were no significant changes between CST1 and 2 for either the CTA or the CS alone groups (Figure 1 – figure supplement 2E-J). Strikingly, however, the neuronal suppression in response to saccharin was significantly reduced in the KO mice; a smaller fraction of neurons were suppressed (Figure 1L), and the average magnitude of suppression decreased (Figure 1J, M). There was a smaller reduction in the population and magnitude of activated neurons as well, but this did not reach statistical significance (Figure 1N, O). These data reveal that Shank3 KO mice have less cortical suppression during tastant delivery and exhibit reduced or slowed acquisition of CTA.

Shank3 knockout increases the correlated variability in AIC during CTA acquisition

We wondered what other features of AIC activity during CTA acquisition might differ between WT and Shank3 KO mice. The correlated variability between neurons – the trial-to-trial co-fluctuation of their activity over repeated presentations of the same stimulus – has been shown to limit the information-encoding capacity of neuronal networks (Moreno-Bote et al., 2014; Bartolo et al., 2020; Rumyantsev et al., 2020; Kafashan et al., 2021; Hazon et al., 2022). We thus wondered whether Shank3 KO might impede CTA by increasing the correlated variability in AIC during tastant presentations. We first characterized the trial-to-trial coactivity between AIC neurons by measuring the pairwise correlations of their mean calcium responses across CS (saccharin) presentations (see methods). There was a rightward shift in the cumulative distribution function of these pairwise correlations in KO compared to WT animals, which was apparent in both conditioning sessions (Figure 2A, B). Comparing the mean pairwise correlation calculated by animal confirmed a persistent increase in the average strength of coactivity in the KO mice across conditioning sessions (Figure 2C).

Shank3 knockout increases the correlated variability in AIC during CTA acquisition

(A, B) Cumulative distribution of taste-elicited coactivity (Pearson coefficients) between neurons during CST1 (A) and CST2 (B). Dashed lines indicate the shuffled data (CST1: WT, n = 6201 pairs, KO, n = 2645 pairs, WT vs. KO, Kolmogorov-Smirnov test, p < 0.0001; CST2: WT, n = 8476 pairs, KO, n = 2847 pairs, WT vs. KO, Kolmogorov-Smirnov test, p < 0.0001). (C) Change in the average taste-elicited coactivity from CST1 to CST2, plotted by animal. The dashed line indicates the chance-level correlation. (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.0142, session: p = 0.4198, interaction: 0.4146). (D) Schematic showing how the signal and noise correlations are computed. (E, F) Change in the signal correlation from CST1 to CST2, plotted by neuronal pair (E) or by animal (F) (E: WT: CST1, n = 6201 pairs, CST2, n = 8476 pairs; KO: CST1, n = 2645 pairs, CST2, n = 2847 pairs, two-way ANOVA, genotype: p = 0.1134, session: p = 0.0009, interaction: 0.5469; F: WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.8143, session: p = 0.4007, interaction: 0.8213). (G, H) Change in the correlated variability from CST1 to CST2, plotted by neuronal pair (G) or by animal (H) (G: two-way ANOVA, genotype: p < 0.0001, session: p = 0.2274, interaction: 0.6112; H: WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.0617, session: p = 0.4017, interaction: 0.7188). (I) Average taste-elicited coactivity in CST1, grouped by inter-neuronal distance and plotted by animal (WT vs. KO, repeated-measures mixed ANOVA, genotype, p = 0.0050, distance, p = 0.7643, interaction: p = 0.5895).

Neuronal coactivity could result from “signal” correlations, which reflect the similarity of average stimulus-tuning properties between neurons, and “noise” correlations that measure trial-to-trial activity co-fluctuations between neurons independent of stimulus tuning (Panzeri et al., 2022). To determine if the “noise” component is the driving force of the increased coactivity observed in the KO mice, we performed a different correlation analysis that allowed us to separate signal-induced correlations from the residual (noise) correlations (Marks and Goard, 2021). For a given neuron, we first obtained the stimulus-induced signal by averaging response traces (3 seconds) across trials (Figure 2D, top). We then estimated the neuron’s trial-to-trial variability (‘noise’) by subtracting this averaged ‘signal’ from individual trials and concatenating the residuals (Figure 2D, bottom). The signal correlation was then quantified by calculating the pairwise correlations of the averaged signal, while the noise correlation was calculated as the pairwise correlations in the trial-to-trial variability. While the signal correlation was reduced from CST1 to CST2, there was no difference between WT and KO mice (Figure 2E, F). By contrast, the correlated variability was significantly higher in the KO mice (Figure 2G, H). The activity correlations between neurons generally decrease when the intraneuronal distances grow wider (Rothschild et al., 2010; Rosenbaum et al., 2017; Khoury et al., 2022). To determine the spatial extent of enhanced correlations in Shank3 KO mice, we measured correlated variability as a function of the inter-somatic distance between neuron pairs (Figure 2I); strikingly, the increased coactivity was observed even in neuronal pairs > 300 μm apart (Figure 2I), indicating a network-wide, rather than local, origin. Thus, in the Shank3 KO mice, there is a widespread increase in correlated variability between AIC neurons, suggesting there is more “noise” in insular cortical circuits.

Finally, to determine if this coactivity is stimulus-dependent, we repeated the correlation analysis using the mean calcium responses two seconds prior to cue onset (Lee et al., 2020). Measuring the trial-to-trial correlations of this spontaneous activity showed that coactivity was enhanced in KO animals even at baseline (Figure 2 – figure supplement 1A-C). Furthermore, as for taste-elicited activity, this increased baseline coactivity had a broad spatial extent (Figure 2 – figure supplement 1D). Altogether, these analyses demonstrate that Shank3 KO mice have increased network-wide, stimulus-independent correlated variability within AIC. This heightened background noise is likely to reduce the signal-to-noise ratio during tastant processing, crucial for the induction of CTA.

Extinction of CTA in Shank3 knockout mice is faster than in wild-type littermates

The reduced expression of CTA in KO mice after the first conditioning session (Figure 1) could be due to a reduction or a delay in the learning process. To differentiate between these possibilities, we measured taste aversion after the second round of conditioning (after CST2), by comparing the ratio of licks to saccharin and water during the first retrieval test session (T1, Figure 1A). We found that both WT and KO mice exhibited strong and comparable CTA during T1 (Figure 3A-D), indicating that CTA formation is slower but still significant in KO mice.

Extinction of CTA in Shank3 knockout mice is faster than in wild-type littermates

(A, B, C) Trial-averaged licks of saccharin (violet) and water (blue) in the CS-only (WT) (A), CTA (WT) (B), and CTA (KO) (C) groups throughout the test sessions (CS-only (WT): repeated-measures ANOVA, taste, p = 0.2530, session, p = 0.4158, interaction, p = 0.4515; CTA (WT): repeated-measures ANOVA, taste, p < 0.0001, session, p < 0.0001, interaction, p < 0.0001; CTA (KO): repeated-measures ANOVA, taste, p = 0.0002, session, p < 0.0001, interaction, p < 0.0001). (D) Trial-averaged lick ratios of saccharin to water throughout the test sessions in CTA (WT) and CTA (KO) groups (WT vs. KO: repeated-measures mixed ANOVA, genotype, p = 0.0069, session, p < 0.0001, interaction, p = 0.0691, post-hoc Tukey test, WT vs. KO, T1 p = 0.8669, T2 p = 0.0032, T3 p = 0.2705, T4 p = 0.9014, T5 p = 0.9531). (E, F) A fitted logistic curve (top) and its derivative (bottom) describing the relationship between lick ratios and test sessions in an example WT mouse (E) and an example KO mouse (F). (G) Average rate of change (derivative) at each test session. The triangle indicates the data mean. (WT vs. KO: repeated-measures mixed ANOVA, genotype, p = 0.0271, session, p < 0.0001, interaction, p = 0.0003; post-hoc Tukey test, WT vs. KO, T1 p < 0.0001, T2 p = 1.0000, T3 p = 1.000, T4 p = 1.000, T5 p = 1.000).

As expected, WT animals underwent gradual memory extinction between T1 and T5 (Berman and Dudai, 2001; Yiannakas and Rosenblum, 2017), with their preference for saccharin slowly returning close to baseline by T4-5 (Figure 3B, D). In contrast, KO mice underwent much faster extinction, with their saccharin consumption returning to baseline by T2 (Figure 3C, D). To calculate the rate of extinction, we fitted a logistic curve to the lick-ratio data for each animal and estimated the rate of memory extinction by taking the derivative of the fit (Figure 3E, F, Kogan and Fontanini, 2024). This analysis revealed distinct extinction dynamics between WT and KO mice (Figure 3E, F): as expected, the lick ratio of saccharin to water increased gradually over time in WT (Figure 3E, top) while rising steeply in KO mice, with a much higher peak rate (Figure 3F, G). Thus, Shank3 KO mice not only form CTA more slowly but also undergo faster CTA extinction than their WT littermates.

CTA learning increases the response reliability of individual AIC neurons

The stimulus responsiveness of individual sensory neurons is prone to drift over time (Deitch et al., 2021; Marks and Goard, 2021; Schoonover et al., 2021). We wondered whether CTA might stabilize taste-elicited insular responses, as has been suggested in other sensory modalities (Poort et al., 2015; Glas et al., 2021; Taylor et al., 2021; Biane et al., 2023; Pancholi et al., 2023). To test this, we identified AIC neurons that could be recorded in each retrieval test session (Figure 4A, see methods) and tracked their daily responsiveness to taste. We first focused on cells that were activated by saccharin in at least one session and found that in the CS-only group, a majority of neurons dropped in and out of the population responses (Figure 4B). In contrast, both WT and KO CTA groups had a higher proportion of neurons reliably activated by saccharin across days (Figure 4C, D). While no neurons in the CS-only group were active across all five sessions, 16 (WT) and 15 (KO) percent of neurons in the CTA conditions were responsive throughout (Figure 4E). The same pattern was found for AIC neurons that were responsive to water, an independent and neutral taste modality that induces distinct neuronal activities in AIC (Figure 4F-I, Katz et al., 2001; Rosen et al., 2010). Hence, associative learning enhances the reliability of AIC neurons’ responsiveness in both genotypes.

CTA learning increases the response reliability of individual AIC neurons

(A) Example images showing neurons present in all five retrieval test sessions (labeled green). Scale bar: 20 μm. (B, C, D) Lifetime plot showing significant responses of the same neurons across test sessions (T1-5) in the CS-only (WT) group (B, n = 42 cells), the CTA (WT) group (C, n = 70 cells), and the CTA (KO) group (D, n = 39 cells). Violet labels activated states, and white labels nonactivated states. (E) Percentage of neurons active for 1, 2, 3, 4, and 5 sessions in (B-D) (Chi-square test, p < 0.0001). (F, G, H) Lifetime plot showing the water-responsiveness of the same neurons across test sessions (T1-5) in the CS-only (WT) group (F, n = 43 cells), the CTA (WT) group (G, n = 63 cells), and the CTA (KO) group (H, n = 36 cells). Blue labels activated states, and white labels nonactivated states. (I) Percentage of neurons active for 1, 2, 3, 4, and 5 sessions in (F-H) (Chi-square test: p < 0.0001).

The stimulus selectivity of individual AIC neurons degrades faster in Shank3 knockout mice during memory extinction

The feature selectivity of individual neurons within primary sensory cortices is dynamic and experience-dependent (Chéreau et al., 2020). We reasoned that taste selectivity – the “breadth of tuning” (Levitan et al., 2019) – of individual AIC neurons might evolve during CTA extinction, and that this process might be accelerated in Shank3 KO mice to promote faster extinction. To test this, we identified neurons activated by water or saccharin delivery, and in test sessions estimated the percentage of neurons that responded to one or both stimuli (Figure 5A). Our analysis revealed a comparable proportion of nonselective AIC neurons (∼40 percent) in both genotypes during the first test session (T1, Figure 5B). During extinction, the percentage of nonselective neurons gradually increased in the WT group and reached ∼50 percent by T5 (Figure 5B); this increase was faster and stronger in KO mice, reaching ∼65 percent by T5 (Figure 5B).

The stimulus selectivity of individual AIC neurons degrades faster in Shank3 knockout mice during CTA extinction

(A) Calcium traces of example neurons activated by saccharin only (left), water only (middle), or both (right). Violet marks the calcium responses to saccharin, and blue marks the responses to water. (B) The percentages of nonselective neurons across test sessions and genotypes. Each data point represents one animal. The colored bar marks the data-averaged value and its standard error. The square indicates the fitted mean in each condition based on the linear mixed-effects model, with session and genotype set as fixed parameters and individual animal as random effect (WT vs. KO, linear mixed-effects model, genotype (in reference to WT): KO, β = −5.146, p = 0.554, session (in reference to T1): T2, β = 5.410, p = 0.438, T3, β = 11.692, p = 0.093, T4, β = 14.433, p = 0.038, T5, β = 8.045, p = 0.248, interaction: KO x T2, β = 16.729, p = 0.103, KO x T3, β = 13.466, p = 0.189, KO x T4, β = −5.420, p = 0.597, KO x T5, β = 21.409, p = 0.037, random effect, variance, 74.187; post hoc t-test, KO, T1 vs. T2, p = 0.0218). (C, D, E) Histograms comparing the distributions of the discriminability index between T1 and T2 (top) and between T1 and T5 (bottom) in the CS-only (WT) group (C, T1: 243 cells, T2: 238 cells, T5: 250 cells, F-test with Bonferroni correction, T1 vs. T2 p = 0.5979, T1 vs. T5 p = 0.2110), the CTA (WT) group (D, T1: 355 cells, T2: 382 cells, T5: 358 cells, F-test with Bonferroni correction, T1 vs. T2 p = 0.6687, T1 vs. T5 p = 0.0001), and the CTA (KO) group (E, T1: 212 cells, T2: 212 cells, T5: 241 cells, F-test with Bonferroni correction, T1 vs. T2 p < 0.0001, T1 vs. T5 p < 0.0001). (F) Comparison of absolute discriminability index across CS-only (WT), CTA (WT), and CTA (KO) groups. Each data point indicates a single-cell value, grouped by animal. Black bars indicate the “predicted” subject-level mean value from the mixed-effects model accounting for both fixed (session, genotype) and random (animal) effects. The colored bars mark the “predicted” marginal means from fixed effects (session and genotype) (linear mixed-effects model, group (in reference to CS-only, T1): CTA (WT), β = 0.1058, p = 0.1538, CTA (KO), β = 0.2442, p = 0.0018; session (in reference to CS-only, T1): T2, β = 0.0271, p = 0.4598, T5, β = −0.0035, p = 0.9239; interaction: CTA (WT) x T2, β = −0.0577, p = 0.2203, CTA (WT) x T5, β = −0.1082, p = 0.0213, CTA (KO) x T2, β = −0.2393, p < 0.0001, CTA (KO) x T5, β = −0.2856, p < 0.0001). Planned post hoc comparisons are tested using linear contrasts on the fitted model with Benjamini-Hochberg correction for multiple comparisons (T2 vs.T1: CTA (WT), p = 0.6455, CTA (KO), p = 0.9742; WT vs. KO: T1, p = 0.6455, T2, p = 0.0301, T3, p = 0.1040).

While binarizing calcium signals into “active” and “non-active” is widely used to describe single-cell stimulus-elicited responses (Zhang and Li, 2018; Gehrlach et al., 2019; Wang et al., 2020; Chen et al., 2021; Staszko et al., 2022; Biane et al., 2023), it ignores changes in response amplitude that could encode stimulus-related information. We therefore performed a complementary analysis and calculated a discriminability index based on the amplitude differences between each neuron’s trial-averaged calcium responses to saccharin and water (see methods, Banerjee et al., 2020; Lee et al., 2021; Kong et al., 2023). A more positive index indicates the neuron has a higher response preference for saccharin, while neurons with a more negative index respond more strongly to water. We first quantified the spread of the distribution by pooling the discriminability indices from all recorded neurons, where a narrowing of the distribution indicates that neurons within the population become more similar in their response preference. In the CS-only group, the variance was similar for the first and last test sessions (Figure 5C). In WT mice that underwent CTA, there was a gradual reduction in the variance (narrowing) of the distribution from T1 to T5 (Figure 5D), and this decrease was already apparent during T2 in KO mice (Figure 5E). Calculating the absolute discriminability index (which reflects the degree of a neuron’s selectivity regardless of the taste stimulus it responds to) revealed that by T5, the mean absolute index was significantly smaller in the CTA group compared to the CS-only group (CS-only (WT) vs. CTA (WT) at T5, p = 0.0213, Figure 5F). This reduction in taste discriminability was again faster and more pronounced in KO mice (CTA (WT) vs. CTA (KO) at T2: p = 0.0301, Figure 5F). This reduction in discriminability was not due to changes in the number and response amplitude of both taste-activated and suppressed AIC neurons, as these remained stable throughout extinction (Figure 5 – figure supplement 1). Thus, although the responses of individual AIC neurons become more reliable after CTA, they gradually provide less information about taste identity during CTA extinction, and this process is accelerated in the KO condition.

Learning improves population coding of taste in AIC, but this enhancement rapidly decays in Shank3 knockout mice

To determine whether the experience-dependent changes in taste discriminability are reflected in population-level activity, we calculated the Euclidean distance between population vectors in response to water or saccharin exposures; this is a widely used measure of the similarity between population neuronal responses to a pair of stimuli (Levitan et al., 2019; Arieli et al., 2022). In the CS-only group, the population responses to water and saccharin remained highly similar, as shown by the consistently small distance between the two (Figure 6A). By comparison, CTA roughly doubled the distance (dissimilarity) between the two taste-elicited population responses, and while this distance decreased during extinction, it remained higher than for the CS-only group (Figure 6A). To determine whether these population differences are sufficient to reliably classify tastants, we trained a support vector classifier to predict the taste class (water or saccharin) based on the ΔF values in the neuronal feature space (see methods, Bishop, 2006; Woods et al., 2020; Raymond et al., 2024). We trained and evaluated a classifier for each animal within each session to capture potential behavioral variability (Twarkowski et al., 2022). While decoding accuracy was close to chance in the CS-only group, CTA learning led to significantly better classification performances across sessions (Figure 6B-C). The distance between population responses to water and saccharin was reduced faster and to a larger degree in KO mice (Figure 6D, linear mixed-effects model: T1 vs. T2 in WT, β = −0.099, p = 0.697; T2 WT vs. KO, β = −0.901, p = 0.015; post hoc paired t-test: T1 vs. T2 in KO, p = 0.009), and this could also be seen in degraded decoding accuracy in these mutant mice (Figure 6E, F). Thus, population-level, discriminative taste representations emerge in AIC following associative learning, but Shank3 KO enhances the decay of this representation during extinction.

Learning improves population coding of taste in AIC, but this enhancement rapidly decays in Shank3 knockout mice

(A) Comparison of the Euclidean distance between population responses to water and saccharin between CTA (WT) and CS-only (WT) groups throughout test sessions (linear mixed-effects model, group (in reference to CS-only): CTA, β = 0.992, p = 0.005, session (in reference to T1): T2, β = −0.065, p = 0.760, T3, β = - 0.046, p = 0.829, T4, β = 0.101, p = 0.636, T5, β = 0.013, p = 0.950, interaction: CTA x T2, β = −0.033, p = 0.910, CTA x T3, β = −0.282, p = 0.336, CTA x T4, β = −0.680, p = 0.020, CTA x T5, β = −0.441, p = 0.132, random effect, variance, 0.312). (B) Diagram showing that training and prediction are performed within each test session. (C) Comparison of decoding accuracies of the linear classifiers trained and tested in each session for the CS-only (WT) and the CTA (WT) groups. The dashed line indicates the chance-level predictions (CS-only (WT) vs. CTA (WT): two-way ANOVA, group, p = 0.0001, session, p = 0.6637, interaction, p = 0.3273). (D, E) Confusion matrices showing the prediction performances in T1 (left) and T5 (right) in an example CTA (WT) mouse (D) and an example CS-only (WT) mouse (E). (F) Comparison of the Euclidean distance between population responses to water and saccharin between CTA (WT) and CTA (KO) groups throughout test sessions (linear mixed-effects model, genotype (in reference to WT): KO β = 0.225, p = 0.625, session (in reference to T1): T2, β = −0.099, p = 0.697, T3, β = −0.328, p = 0.195, T4, β = −0.578, p = 0.022, T5, β = −0.427, p = 0.091, interaction: KO x T2, β = −0.901, p = 0.015, KO x T3, β = −0.621, p = 0.094, KO x T4, β = - 0.674, p = 0.069, KO x T5, β = −0.618, p = 0.095, random effect, variance, 0.538; post hoc t-test, KO, T1 vs. T2, p = 0.0089). (G) Comparison of decoding accuracies of the linear classifiers trained and tested in each session for the CTA (WT) and the CTA (KO) groups. (WT vs. KO: two-way ANOVA, genotype, p = 0.0078, session, p = 0.7293, interaction, 0.7757). (H) Confusion matrices showing the prediction performances in T1 (left) and T5 (right) in an example CTA (KO) mouse. (I) Heatmaps of decoding accuracies using the classifiers trained on T1 responses and tested across sessions. Each row indicates an animal. (J) Quantification of the decoding accuracies in (I), comparing CTA (WT) and CTA (KO) groups (linear mixed-effects model, genotype (in reference to WT): KO, β = −1.417, p = 0.846, session (in reference to T1): T2, β = 0.031, p = 0.995, T3, β = −6.286, p = 0.214, T4, β = −7.545, p = 0.136, T5, β = - 10.031, p = 0.047, interaction: KO x T2, β = −17.732, p = 0.017, KO x T3, β = −14.260, p = 0.054, KO x T4, β = −8.013, p = 0.279, KO x T5, β = −8.261, p = 0.264, random effect, variance, 96.479; post hoc t-test, KO, T1 vs. T2, p = 0.0063).

Lastly, we examined the stability of taste representations in AIC during memory extinction (across sessions) by using the classifiers trained on T1 responses to predict taste classes in the subsequent sessions (T2-5). In the CS-only group, the classifier performed poorly across sessions (Figure 6G). In contrast, in the CTA (WT) condition, the cross-session classifier performed well during the early stages of memory extinction, and the accuracy gradually decreased (T1 vs. T5, p = 0.037, Figure 6G, H). Finally, consistent with our previous findings, in the CTA (KO) condition, the decoding accuracy declined sharply from T1 to T2 (Figure 6G, H). In summary, taste representations in AIC become and remain discriminative after learning. In WT animals, these representations continue to diverge from those during the early memory retrieval as CTA is extinguished, and in KO animals, this shift is faster.

Discussion

Flexible sensory coding ensures that perceptions are updated to reflect their value to an organism, to enable adaptive behaviors. While sensory coding deficits have been reported in mouse models of ASD, including Shank3 KO (Monday et al., 2023), it is unknown whether flexible coding is impaired in ethological learning paradigms, such as CTA. Here, we used 2-photon calcium imaging and a head-restrained CTA learning paradigm to address this question in Shank3 WT and KO mice. We show that neuronal ensembles in AIC, the cortical region critical for taste processing, have higher correlated noise and lower taste discriminability in Shank3 KO compared to WT littermates, and these changes are correlated with slower CTA learning, indicating that Shank3 KO animals require repeated exposure to aversive experiences to enable updating of taste valence. Once CTA is established, responses of AIC neurons in both wild-type and Shank3 KO mice are initially able to encode and discriminate between different tastants, both at the single-cell and population levels. However, during repeated exposure to the conditioned tastant, taste discriminability in Shank3 KO mice degrades faster than in their wild-type counterparts, and this coincides with the premature extinction of CTA. This work demonstrates that loss of Shank3 compromises the flexibility of sensory coding, and thus the ability of these animals to adapt their behavior in response to negative outcomes.

The increased correlated variability between AIC neurons observed in Shank3 KO mice during CTA learning may impede information encoding (Moreno-Bote et al., 2014; Bartolo et al., 2020; Rumyantsev et al., 2020; Kafashan et al., 2021; Hazon et al., 2022) and contribute to the slower acquisition of CTA. Reduced cortical feedback inhibition has been shown to increase the correlated variability (noise correlation) in the cortex (Stringer et al., 2016). Indeed, we detected reduced neuronal suppression in the AIC of Shank3 KO mice, and because Shank3 deletion impairs cortical interneuron functions (Lee et al., 2017; Chen et al., 2020; Pagano et al., 2023), the observed heightening of correlated variability could stem from weakened local inhibition within the AIC network. Heightened noise correlation has been observed across sensory cortices in different animal models of ASD, including the somatosensory and visual cortices of mice with cell-type-specific deletion of Syngap1 and MeCP2, respectively (Banerjee et al., 2016; Zhao and Kwon, 2023), but the spatial extent of the noise correlations in these mutant animals has not been examined. Because the increased correlated variability in Shank3 KO mice occurs over a large spatial scale, it could reflect stimulus-independent, correlated activity of divergent inputs to AIC. Regardless of the source, this increased correlated noise is likely to degrade taste-elicited responses and thus hinder CTA encoding. Our behavioral results are also consistent with a recent study showing a slower acquisition of CTA in a pharmacologically induced rat model of autism (Pal et al., 2025). Taken together with our finding that excessive cortical noise correlations impede discriminative sensory learning in Shank3 KO mice, these observations suggest that a reduced signal-to-noise ratio driven by high correlated variability in the sensory cortex is a common mechanism underlying maladaptive sensory processing in ASD.

AIC neurons exhibit a wide range of tuning properties (Fletcher et al., 2017; Levitan et al., 2019; Chen et al., 2021; Staszko et al., 2022), and we found that, similar to neurons in the primary visual and somatosensory cortices (Poort et al., 2015; Khan et al., 2018; Chéreau et al., 2020), the stimulus selectivity (response preference) in AIC can be dynamically adjusted at the cellular level, from being selective to nonselective during CTA extinction. While CTA learning can still initiate selective response preferences in AIC neurons of Shank3 KO mice, this stimulus selectivity was rapidly degraded during the premature extinction of CTA memory. The maintenance and extinction of a faithful CTA memory rely on both Hebbian and homeostatic synaptic plasticity mechanisms within AIC (Escobar et al., 1998; Li et al., 2016; Rodríguez-Durán et al., 2017; Haley et al., 2020; Wu et al., 2021). Because Shank3 KO is known to perturb several of these critical plasticity processes (Bozdagi et al., 2010; Raynaud et al., 2013; Tatavarty et al., 2020), rapidly reduced stimulus selectivity of AIC neurons could result from defects in one or more of these local plasticity mechanisms. An alternative, though not mutually exclusive, possibility is that the rapid loss of stimulus selectivity and CTA memory results from the impairment in other brain regions that project to AIC, such as BLA and the ventral tegmental area (VTA) (Levitan et al., 2020; Gil-Lievana et al., 2022). Hebbian plasticity can be induced at synaptic inputs from BLA to AIC during CTA learning (Escobar et al., 1998; Rodríguez-Durán et al., 2017; Haley et al., 2020), and dopaminergic projections from VTA facilitate the induction of NMDAR-dependent LTP within AIC (Rodríguez-Durán et al., 2023). Thus, Shank3 KO could disrupt function in these upstream brain regions to impair plastic changes in AIC during extinction. This possibility is consistent with the finding that Shank3 down-regulation in VTA accelerates extinction of socially induced conditioned place preference (CPP) memory (Bariselli et al., 2018).

During CTA extinction, population responses to the neutral and conditioned taste stimuli in AIC gradually become more similar. Because CTA learning can be suppressed by the familiarity of a tastant (De la Casa and Lubow, 1995), we were not able to compare the population responses to different taste stimuli before and after CTA learning; however, even after extinction these population responses were more distinct in the CTA than the CS-only group (demonstrated by the Euclidean distance and the classifier prediction), suggesting that taste coding in AIC does not revert to its naïve state even when the behavioral aversion is gone. This is consistent with previous in vivo electrophysiology demonstrating that, during CTA extinction, the population-level taste coding in AIC continuously evolves but does not return to the pre-CTA state (Moran and Katz, 2014). In contrast, a recent study (Raymond et al., 2024) reported that the population responses to conditioned and appetitive taste stimuli converge to a much greater degree than we observed after CTA extinction. One explanation for the difference between Raymond et al. and our findings is that we used two conditioning sessions with a novel tastant, while Raymond et al. used a single conditioning session with a familiar tastant. Thus, our paradigm likely elicited stronger CTA, and thus a more enduring ability to discriminate tastants. Interestingly, in contrast to the WT condition, extinction in the Shank3 KO leads to a rapid and stronger convergence of population taste-evoked responses in AIC, which ultimately reduces the ability to decode the identity of taste stimuli from their responses. Degraded sensory coding has been observed in sensory cortices of other ASD mouse models (Monday et al., 2023; Wang and Feldman, 2024); thus, unstable and poorly differentiated cortical responsivity to sensory stimuli is a pathogenic feature shared across brain regions and gene mutations in ASD.

Our finding that taste representations remain discriminative in AIC throughout CTA extinction raises an important question: what taste information does AIC population activity actually encode? In the hippocampus, representations of odor “identity” remain stable independent of learning, while those encoding odor “values” are susceptible to extinction (Biane et al., 2023). In AIC, ensemble activity encodes both the identity and palatability of taste stimuli, as well as related decision-making, in sequential temporal epochs (Katz et al., 2001; Sadacca et al., 2016; Livneh et al., 2017; Lavi et al., 2018; Chen et al., 2021; Staszko et al., 2022; Kogan and Fontanini, 2024; Raymond et al., 2024). The population AIC activity we analyzed (3-second duration after delivery) is unlikely to simply encode taste identity, because if that were the case, the defined taste representations would be independent of learning and should be present even in animals that didn’t receive LiCl. Instead, the AIC population activity we measure likely encodes primarily taste palatability, or “valence.” This possibility would also explain the poor decoding performances of classifiers trained in the CS-only condition, where water and saccharin share similar innate palatability (Staszko et al., 2022).

Flexible processing of new sensory information in the cortex enables animals to update their perceptions and modify behaviors that are critical for survival, and disruptions in this process could contribute to maladaptive behaviors in neurodevelopmental disorders. Here, we show that a higher signal-to-noise ratio in the AIC of Shank3 KO mice may hinder the updating of taste valence following negative experiences, leading to slower CTA learning and faster extinction. This rapid extinction is associated with rapid loss of taste discriminability within AIC, and this response instability ultimately compromises the ability to decode taste identity from AIC responses. More broadly, these data suggest that behavioral inflexibility in some forms of ASD could arise from a reduced ability to update behavior when the valence of a sensory input changes.

Materials and methods

Key Resources Table

Resource Availability

Lead contact and materials availability

Further information and requests for resources and reagents should be directed to the lead contact, Gina G. Turrigiano (turrigiano@brandeis.edu).

Experimental Model and Subject Details

All experimental procedures were approved by Brandeis University Institutional Animal Care and Use Committee and followed the National Institute of Health guidelines for the Care and Use of Laboratory Animals. C57BL/6 and Shank3B knockout (Δexon 13-16, JAX: 017688) mouse strains were obtained from the Jackson Laboratory, and the colonies were maintained in Foster Biomedical Research Labs at Brandeis University. Adult mice (8-16 weeks old) were used in experiments performed in this study. All animals were grouped with their littermates or housed individually once they entered the behavioral training paradigm. In all cases, animals were housed in a humidity- and temperature-controlled environment and entrained to a 12-hour light-dark cycle (light phase from 7:00-7:00), with ad libitum access to food and water unless described otherwise. Mice of each sex were assigned equally to each experimental condition. In this paper, 7, 9, and 9 animals were assigned to the CS-only (WT), CTA (WT), and CTA (KO) groups in the two-conditioning behavior experiments (Figures 1 and 3). For calcium imaging analysis, 1 and 2 animals were excluded from the CTA (WT) and CTA (KO) groups, respectively, due to a lack of sufficient GCaMP expression (Figures 1-6). In the separate single-conditioning paradigm, 13 and 13 animals were used in the CTA (WT) and CTA (KO) groups.

Viral Vectors

The AAV serotype 1 expressing pGP-AAV-syn-jGCaMP8s (Addgene: 162374) was obtained from Addgene.

Stereotaxic Surgery

Mice (8-10 weeks old) were anesthetized with a constant airflow containing 2% isoflurane throughout the procedure. The skull was exposed, and a craniotomy was created above the right AIC. AAV expressing GCaMP8s (titer: ∼1.2 x 1013, 1:3 dilution, 800 nL in total) was slowly infused into AIC (anterior-posterior (AP) with reference to bregma: 1.0 mm, medial-lateral (ML): - 3.8 mm, dorsal-ventral (DV): with reference to the brain surface: −2.0 mm) through a micropipette connected to a microinjector (Narishige, MO-10), at a rate of approximately 200 nL/min. To allow the virus particles to spread, the pipette was kept in place for another 5 minutes before being slowly withdrawn from the site. Two weeks after the virus injection, the mice (10-12 weeks old) then underwent the implantation of a gradient-index (GRIN) lens and a customized titanium head bar. Mice were anesthetized as described above, and a dose of dexamethasone (2 mg/kg) was administered subcutaneously at the beginning of the surgery to minimize brain swelling. Once the skull was revealed, a craniotomy slightly larger than 0.5 mm x 0.5 mm was made at the site of the previous virus injection. To create a leading track, a 25-gauge blunt needle was slowly lowered until it reached above AIC (AP: + 1.0 mm, ML: −3.8 mm, DV: −1.8 mm), remained in place for five minutes and was withdrawn from the site. A GRIN lens (diameter: 0.5 mm, length: ∼4 mm, CLHS050GFT126, GoFoton) was then gradually lowered through a customized holder until its bottom surface rested on top of AIC (AP: + 1.0 mm, ML: −3.8 mm, DV: −2,0 mm). The lens was secured to the skull by Vetabond and the UV-curable adhesive, after which a titanium head-bar was installed. Metabond tinged with black ink was applied to further cement the lens and the head bar while covering and sealing the skull. After the operation, animals were kept and monitored individually in their home cages, and a daily dose of meloxicam (1-2 mg/kg) was given subcutaneously for at least three days to reduce brain inflammation. Animals were allowed for at least one week of recovery before starting behavioral training.

Head-restrained CTA Learning Paradigm

One day before the mice (13-15 weeks old) began training in the head-fixed position, ad libitum water access was removed from their home cages to motivate their engagement in the subsequent tasks. The restricted water access was then given daily through the motorized spout during the training task or in the home cage for thirty minutes (described in detail below). Each training day, animals were placed on a customized wheel (Livneh et al., 2017) with their heads restrained and underwent a thirty-minute training session consisting of 30-35 trials. In each trial, a two-second monotonal acoustic cue (2 kHz) was initiated to signal the onset of the trial, with a motorized spout simultaneously moving to the location close to the animal’s mouth. Mice then learned to perform five dry licks in order to receive the liquid through the spout. The gravity-based liquid delivery, once initiated, lasted 3 seconds, after which the spout was moved away from the animal. A 10-second intertrial rest followed before the next trial started. A trial was considered failed if the animal did not reach the threshold to initiate the liquid delivery within sixty seconds after the spout was in place.

Habituation

During the habituation stage, water was used as the tastant. If animals refused to lick the spout throughout the session or their body weight dropped over 15 percent of the initial weight, an additional ad libitum access to water was granted for 30 minutes in the home cage. On average, it took 10 days for the animals to become familiarized with the task, reaching consistent performance without failed trials within thirty minutes. By the end of the habituation, the spout became the only daily source of liquid the animals received, and their body weight stabilized.

Conditioning

Once habituated, animals then received two conditioning sessions to induce the formation of conditioned taste aversion (CTA). On each conditioning day, animals underwent the training session with the same scheme as in habituation, except that saccharin (25 mM, CS) was given instead of water. Five minutes after the session, a dose of lithium chloride (LiCl, 0.15 mM, 2% body weight, US) was intraperitoneally injected to induce gastric malaise. Afterward, animals remained in place for another 25 minutes before returning to the home cage. To overcome the potential association between malaise and the context, which might confound animals’ subsequent performance, animals were allowed to receive water in a recovery session (in the same environment) the next day after each conditioning; during this time, animals usually resumed their engagement in the task.

CTA retrieval and extinction testing

After the second recovery session, we subjected animals to consecutive test sessions for five days to measure the retrieval and extinction of CTA memory. In each test session, water was given for the first three trials to engage animals in the task. Afterward, water and saccharin were delivered in a semi-randomized order such that animals could not predict the identity of the tastant in the upcoming trial while overall receiving similar numbers of trials that delivered water and saccharin.

Single-conditioning CTA paradigm

A modified taste learning paradigm was conducted to examine whether Shank3 knockout mice expressed taste aversion when presented with choices after one conditioning. After the first conditioning session and the subsequent recovery session, animals were immediately subjected to a retrieval test session in which water and saccharin at various concentrations (25 mM, 10 mM, 5 mM, 1 mM) were delivered in a semi-randomized order. Trial-averaged licks and lick ratios were quantified as described below.

Behavior Data Analysis

Trial-averaged lick calculation

Throughout the paradigm, licks were detected by a capacitive sensor (MPR121, Adafruit) attached to the spout, which generated TTL signals recorded with the customized code and time-stamped to the frames during two-photon image acquisition. To obtain trial-averaged licks of a tastant (Figure 1 and 3), total lick counts during the 3-second delivery were averaged across trials delivering that tastant in a given session. In the cases where the licks of saccharin and water were compared (Figure 3), the ratio was further calculated by dividing the trial-averaged licks of saccharin over that of water.

Logistic curve fitting

A logistic curve was fitted to the trial-averaged lick ratios as a function of time (session) to describe the temporal dynamics of the licking behavior (Kogan and Fontanini, 2024). The logistic function was fitted using the curve_fit function in the Python scipy model and is defined as: where f(t) is the fitted lick ratio at time t; L is the maximum lick ratio; k denotes the growth rate, and t0 indicates the inflection point. The derivative of the logistic function was calculated as:

Immunohistochemistry

Animals were deeply anesthetized with isoflurane and perfused with 4% paraformaldehyde (PFA). The whole brain was extracted, post-fixed in 4% PFA for 24 hours, and sliced on a vibrating microtome (Leica Vibratome VT 1200s). Coronal brain slices (50 µm) containing AIC were collected; to verify the placement of the GRIN lens, the slices with the visible track left by the lens were selected, washed three times with PBS, permeabilized with the blocking buffer (5% goat serum/3% BSA/0.3% Triton X-100 in PBS) at room temperature for 2 hours, and incubated with primary antibodies diluted in the blocking buffer at 4°C overnight. To confirm the expression of GCaMP8s, chicken anti-GFP (1:1000, Aves Labs) and mouse anti-NeuN (1:500, MAB-377, Millipore) were used. On the next day, the slices were washed three times with PBS and incubated with secondary antibodies diluted in the buffer containing 5% goat serum and 3% BSA at room temperature for 3 hours (goat anti-chicken Alexa-488 and goat anti-mouse Alexa-594, 1:400, Thermo-Fisher). After another three washes with PBS, the slices were mounted on the slides using the DAPI-Fluoromount-G mounting medium (SouthernBiotech), cover-slipped, and stored at 4°C until imaging.

Confocal Imaging

To verify the expression of GCaMP8s and the placement of the GRIN lens, stained slices were imaged on a laser scanning confocal microscope (Zeiss LSM880) using the tiling scan under a 20x objective, with a frame size of 1024 pixels x 1024 pixels. For each hemisphere, the images were stitched online using the ZEN Black software (Zeiss) and then analyzed using ImageJ/FIJI (NIH, US).

Two-photon Imaging

Acquisition

Two-photon imaging was conducted using a resonant-scanning two-photon microscope (Neurolabware, 31 frames per second; 512 pixels x 512 pixels) with a 10X objective (Thorlabs, NA: 0.50, air, working distance: 7.77 mm, zoom factor: 2.8). GCaMP8s was excited at 920 nm using a tunable ultrafast laser (SpectraPhysics Insight X3, 50-80 mW measured at the front aperture of the objective). Fields of view of all imaging were selected approximately 200 µm below the bottom surface of the GRIN lens. On average, 750 frames (25 seconds) were obtained for each trial.

Processing and registration

For each imaging session, frames from all successful trials were included and concatenated into a series. In each movie, the movement was corrected, the regions of interest (ROI) corresponding to individual neurons in the field of view were identified, and the raw fluorescence from the cell bodies and neuropils was extracted using Suite2P (Pachitariu et al., 2017). The processed images were then inspected, and if needed, the ROIs were manually selected or excluded using the Suite2P GUI so that only the cell bodies were used for further analysis. Only the ROIs near the center of the field of view were included to avoid any potential peripheral distortion of fluorescence by spherical aberration (Sattin et al., 2024). The time series of neuropil-corrected calcium signals of individual neurons were obtained by subtracting the time series of neuropil fluorescence (scaled by 0.7) from that of the raw fluorescence of cell bodies. In the case where multi-day cell registration was required (i.e., identification of the same cells across sessions, Figure 4 and 6), Spatial footprints of each neuron were extracted from the field of view of each session with in-house customized code and were aligned across sessions using the probabilistic modeling method implemented in CellReg (Sheintuch et al., 2017). Centroid distance was recommended by CellReg in most cases, and a probability threshold of 0.5 was used.

Cell inclusion criteria

Cells were identified by the classifier built and trained in Suite2p. Once identified, each cell’s extracted somatic and neuropil calcium traces were then inspected; if the contamination from neuropils was severe (i.e., the magnitude of neuropil calcium signals was consistently higher than that of somatic signals), that cell was excluded from the analysis.

Calcium Imaging Data Analysis

Calculation of ΔF/F0 and stimulus responsiveness

Z-scored changes in calcium signals were used to represent the temporal activity dynamics of individual neurons in all analyses. In each trial, neuropil-corrected calcium traces of the same duration (total 656 frames) were first extracted, which consisted of 155 frames of baseline (5 seconds before the cue onset), 62 frames of the acoustic cue (2 seconds), 15 frames of dry licks (∼ 0.5 seconds), 112 frames of tastant delivery (3.5 seconds), and 312 frames of post-delivery rest (10 seconds). On average, it took animals no more than 0.5 seconds (15 frames) to reach the threshold of five dry licks; however, during the early stages of memory extinction, animals occasionally hesitated or even failed to participate in some trials, prolonging the dry-lick period to varying degrees. In such cases, we only included 15 frames starting from the initiation of the first dry lick to ensure that the lengths of the time series were consistent across animals, trials, and sessions. For each neuron in a given trial, we first computed its changes in fluorescence (ΔF/F0) by subtracting the mean fluorescence of baseline (F0) from the fluorescence at each time point and dividing the result values by F0. Next, we calculated the z-scores using the formula (F(t) – Fbaseline) / SD, where F(t) represents the ΔF/F0 at time t, and Fbaseline and SD denote the mean and standard deviation of the ΔF/F0 values during the baseline. To determine whether a neuron is responsive to a taste stimulus in a given session (Figure 1, 2, 4, 5), we compared the mean z-scored activity during the taste delivery (last 2 seconds) to the mean z-scored baseline activity (2 seconds before the cue onset) across trials using paired t-test. A neuron was categorized as “activated” and “suppressed” during the tastant delivery if there was a significant increase and decrease in activity compared to the baseline, respectively. We omitted the first second of the tastant delivery when calculating the mean activity to avoid a potential carryover from the dry lick responses (Chen et al., 2021). In the analyses that used the amplitude of the trial-averaged response (z-scored ΔF/F0, Figure 1, 2, 4, 5, 6), we first averaged the z-scored calcium traces across trials and then calculated the values of the mean activity during tastant deliveries as described above.

Coactivity analysis

The calculation of taste-elicited coactivity patterns between neuronal pairs was adapted from Twarkowski et al., 2022. Within each session, trials of the same tastant deliveries were selected, and the Pearson correlation coefficient between two neurons’ mean z-scored ΔF/F0 values during the tastant delivery across these trials was calculated. To determine the stimulus-independent baseline coactivity, the mean z-scored ΔF/F0 values during the 2-second period before the cue onset were used to compute the Pearson correlation coefficient.

Pairwise signal and noise correlations

The signal and noise correlations were calculated as previously described (Marks and Goard, 2021). The signal correlation was obtained in a given session by calculating the Pearson correlation between two neurons’ full trial-averaged activity traces during the tastant delivery. To estimate the correlated variability (noise), we first subtracted a neuron’s trial-averaged activity trace from its z-scored activity trace in each trial. The subtracted traces (residuals) from all trials within a session were then concatenated into a whole series representing trial-to-trial variability in the neuron’s activities. The Pearson correlation coefficient was calculated between two neurons’ concatenated, subtracted traces to represent correlated variability.

Single-cell discriminability index

The discriminability index was adapted from the previously reported method (Kong et al., 2023) and indicated a neuron’s response preference between taste stimuli (water or saccharin). In a given session, the discriminability index for each neuron is defined as: where µ1 and µ2 denote the trial-averaged activity amplitude averaged across saccharin and water trials, respectively, and the denominator is the pooled standard deviation. A positive index indicates that the neuron responds more preferentially to saccharin, while a more negative index indicates a higher response preference for water.

Euclidean distance

We calculated the pairwise Euclidean distance to estimate the similarity between population responses to saccharin and water (Levitan et al., 2019; Chen et al., 2021; Raymond et al., 2024). In a given test session, for each animal, we constructed the trial-averaged ΔF values of all neurons into two population vectors, one responding to saccharin and the other to water. The Euclidean distance between these two vectors is then defined as: Where v1 and v2 denote the population vector in response to saccharin and water, respectively, and n is the number of neurons included in the vectors. To mitigate the influence on Euclidean distance by the number of neurons included in the calculation (i.e., the dimension), we normalized the distance by the square root of the dimension (Levitan et al., 2019).

Decoding analysis

We trained a support vector classifier for each animal to predict the tastant class (water or saccharin) in a given session, using the SVC function with a linear kernel in the Python scikit-learn module (Woods et al., 2020; Twarkowski et al., 2022; Raymond et al., 2024). We first constructed population response vectors for each trial (labeled as saccharin or water), and these vectors constituted the dataset for model training and testing. The linear decoder is defined as: where L(t) is the predicted tastant class, and v(t) is the population response vector for the trial t. θ(z) is the Heaviside step function that ensures hard classification, while w denotes the vector that contains weights assigned to each neuron, and b is the bias term (Bishop, 2006). We then used 10-fold cross-validation, a standard method for model validation and evaluation, to train and test the data. Cross-validation produced an averaged accuracy score that indicated the classifier’s performance. To calculate chance-level prediction, we shuffled the labels and trained the classifier as described above for 1000 iterations. This shuffling process resulted in an average 50-percent accuracy score by chance. To test the stability of insular taste representations, we used the classifier trained on the data in test 1 to predict the tastant classes in the subsequent sessions and calculated the accuracy score accordingly.

Quantification and Statistical Analysis

All statistical analyses were performed using the Python scipy, statsmodels, or pingouin packages. To assess the main effects and interactions in experiments comparing the same animals or neurons across sessions under different conditions, a paired t-test or repeated-measures mixed ANOVA was used unless stated otherwise. For experiments requiring specific comparisons between conditions, a linear mixed-effect model was fitted to the data, and the relevant coefficient was extracted. If the comparison could not be directly inferred from the coefficient, a post hoc t-test was performed. The significance levels were labeled by asterisks (*): *: P < 0.05; **: P < 0.01; ***: P < 0.001; ****: P < 0.0001.

eLife Assessment

This valuable study provides solid evidence for deficits in aversive taste learning and taste coding in a mouse model of autism spectrum disorders. Specifically, the authors found that Shank3 knockout mice exhibit behavioral deficits in learning and extinction of conditioned taste aversion, and calcium imaging of the gustatory cortex identified impaired neuronal responses to taste stimuli. This paper will likely be of interest to researchers studying how learning and sensory processes are affected by genetic causes of autism spectrum disorders.

Reviewer #1 (Public review):

Summary:

The study from Wu and Turrigiano investigates how disruption of taste coding in a mouse model of autism spectrum disorders (ASDs) affects aversive learning in the context of a conditioned taste aversion (CTA) paradigm. The experiments combine 2-photon calcium imaging of neurons in the gustatory portion of the anterior insular cortex (i.e., gustatory cortex) with behavioral training and testing. The authors rely on Shank3 knockout mice as a model for ASDs. The authors found that Shank3 mice learn CTA more slowly and extinguish the memory more rapidly than control subjects. Calcium imaging identified impairments in taste-evoked activity associated with memory encoding and extinction. During memory encoding, the authors found less suppressed neuronal activity and increased correlated variability in Shank3 mice compared to controls. During extinction, they observed a faster loss of taste selectivity and degradation of taste discriminability in mutants compared to controls.

Strengths:

This is a well-written manuscript that presents interesting findings. The results on the learning and extinction deficits in Shank3 mice are of particular interest. Analyses of neural activity are well conducted and provide important information on the type of impaired cortical activity that may correlate with behavioral deficits.

Weaknesses:

(1) The experiments rely on three groups: CS-only WT, CTA WT, and CTA KO. Can the authors provide a rationale for not having a CS-only KO group?

(2) The authors design an effective behavioral paradigm comparing consumption of water and saccharin and tracking extinction (Figure 3). This paradigm shows differences in licking across distinct behavioral conditions. For instance, during T1, licking to water strongly differs from licking to saccharin for both WT and KO. During T2, licking to water strongly differs from licking to saccharin only for WT (much less for KO), and licking to saccharin in WT differs from that in KO. These differences in taste sampling across conditions could contribute to some of the effects on neural activity and discriminability reported in Figures 5 and 6. That is sucrose and water trials may be highly discriminable because in one case the mouse licks and in the other it does not (or licks much less). The author may want to address this issue.

(3) Are there any omission trials following CTA? If so, they should be quantified and reported. How are the omission trials treated with regard to the analyses?

(4) The authors describe the extinction paradigm as "alternative choice". In decision-making, alternative choice paradigms typically require 2 lateral spouts to report decisions following the sampling from a central spout. To avoid confusion, the authors may want to define their paradigm as alternative sampling.

(5) Figure 4 reports that CTA increases the proportion of neurons that consistently respond to saccharin and water across days. While the saccharin result could be an effect of aversive learning, it is less clear why the phenomenon would generalize to water as well. Can the authors provide an explanation?

(6) The recordings are performed in the part of the anterior insular cortex that is typically defined as "gustatory cortex" (GC). Given the functional heterogeneity of the anterior insular cortex (AIC) and given that the authors do not sample all of the anteroposterior extent of AIC, I would suggest being more explicit about their positioning in GC. Also, some citations (e.g., Gogolla et al, 2014) refer to the posterior insular cortex, which is considered more inherently multimodal than GC. GC multimodality is typically associative in nature, as only a few neurons respond to sound and light in naïve animals.

(7) It would be useful to add summary figures showing the extent of viral spread as well as GRIN lens placement.

(8) I encourage the authors to add Ns every time percentages are reported. How many neurons have been recorded in each condition? Can the authors provide the average number of neurons recorded per session and per animal?

(9) It looks like some animals learned more than others (Figure 1E or Figure 3C). Is it possible to compare neural activity across animals that showed different degrees of learning?

Reviewer #2 (Public review):

Wu and Turrigiano investigated how cortical taste coding during conditioned taste aversion (CTA) learning is affected in Shank3 knockout (KO) mice, a model of monogenic ASD. Using longitudinal two-photon calcium imaging of AIC neurons, the authors show that Shank3 KO mice exhibit reduced suppression of activity in a subset of neurons and a higher correlated variability in neural activity. This is accompanied by slower learning and faster extinction of aversive taste memories. These results suggest that Shank3 loss compromises the flexibility and stability of cortical representations underlying adaptive behaviour.

Major strengths:

(1) Conceptual significance: The study connects a molecular ASD risk gene (Shank3) to flexible sensory encoding, bridging genetics, systems neuroscience, and behaviour.

(2) Technical rigour: Longitudinal calcium imaging with cell-registration across learning and extinction sessions is technically demanding and well-executed.

(3) Behavioural paradigm: The use of both acquisition and extinction paradigms provides a more nuanced picture of learning dynamics.

(4) Analyses: Correlated variability, discriminability indices, and population decoding analyses are robust and appropriate for addressing behavioural and network-level coding changes.

Major weaknesses:

(1) Causality: The paper infers that increased correlated variability causes learning deficits, but no causal tests (e.g., optogenetic modulation of inhibition or interneuron rescue) are presented to confirm this.

(2) Behavioural scope: The study focuses exclusively on taste aversion; generalisation to other flexible learning paradigms (e.g., reversal or probabilistic tasks) is not addressed.

(3) Mechanistic insights: While providing interesting findings of altered sensory perception and extinction of learning-related signals in AIC, it offered nearly no mechanistic insights. This makes the interpretation, especially on how generalisable these findings are, difficult. Also, different reported findings are "potentially" connected, but the exact relation between increased correlated variability and faster loss of taste selectivity cannot be assessed.

Reviewer #3 (Public review):

In this study, Wu & Turrigiano investigate an ethologically relevant form of associative learning (conditioned taste aversion - CTA) and its extinction in the Shank3 KO mouse model of ASD. They also examine the underlying circuits in the anterior insular cortex (AIC) simultaneously, using two-photon calcium imaging through a GRIN lens. They report that Shank3 KO mice learn CTA slower and suggest that this is mediated by a reduction in tastant-stimulus activity suppression of AIC neurons and a reduced signal-to-noise ratio due to increased noise correlations in AIC neurons. Interestingly, once Shank3 KO mice acquire CTA, they extinguish the aversive memory more rapidly than wild-type mice. This accelerated extinction is accompanied by a faster loss of neuronal and population-level taste selectivity and coding in the AIC compared to WT mice.

This is an important study that uses in vivo methods to assess circuit dysfunction in a mouse model of ASD, related to sensory perception valence (in this case, taste). The study is well executed, the data are of high quality, and the analytical procedures are detailed. Furthermore, the behavioural paradigm is well thought out, particularly the approach for assessing extinction through repeated retrieval sessions (T1-T5), which effectively tests discrimination between saccharin and water rather than relying solely on lick counts or total consumption as a measure of extinction. Finally, the statistical tests used are appropriate and justified.

There is, however, a missing link between the behavioural findings and the underlying mechanisms. More specifically:

(1) The authors don't make a causal link between the behaviour and AIC neurophysiology, both the percentage of suppressed cells and the coactivity measurements. For the % of suppressed cells, it seems that both WT and KO cells are suppressed in the transition between CST1 and CST2 (Figure 1L), yet only the WT mice exhibit CTA (at least by CST2). For the taste-elicited coactivity measure, it seems that there is an increase in coactivity from CST1 to CST2 in WT (Figure 2C - blue, although not statistically tested?), but persistently higher coactivity in KO. Is this change of coactivity in WT important for the expression of CTA? Plotting behavioral performance (from Figure 1G) against coactivity (from Figure 2C) for each animal would be informative.

(2) Shank3 KO cells already show an increase in baseline coactivity (Figure 2- figure supplement 1), and the authors never examine CS-only responses in the KO group, therefore making it difficult to determine whether elevated coactivity and noise correlations reflect a generalized AIC abnormality in Shank3 KOs (perhaps through impaired PV-mediated inhibition in insular cortex - Gogolla et al, 2014) that is not directly responsible/related to CTA?

(3) How do the authors interpret the large range of lick ratios (Figure 1G) for WT (almost bi-modal distribution)? Is there a within-subject correlation with any of the neurophysiological measurements to suggest a relationship between AIC neurophysiology and behavioural expression of CTA?

(4) Indeed, CTA appears to be successfully achieved for Shank3 KO mice delayed by 1 day, as the level of saccharin aversion during the first retrieval session (T1) is comparable between Shank3 KO and WTs. In this context, not extending the first part of the paradigm to include CST3 seems to be a missed opportunity. Doing so would have allowed for within-cell and within-subject comparison of taste-elicited pairwise correlation across the learning and to investigate the neural mechanism of delayed extinction in KOs more effectively.

(5) How to interpret Figure 5F: Absolute discriminability is lower for T5 for CTA WT and CTA KO compared to CS-only? Why would AIC neurons have less information on taste identity by the end of extinction than during the unconditioned (CS-only) condition? And if that is the case, how is decoding accuracy in Figure 6C higher in T5 for CTA WT vs CS-only?

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

(1) The experiments rely on three groups: CS-only WT, CTA WT, and CTA KO. Can the authors provide a rationale for not having a CS-only KO group?

We did not include the CS-only (KO) group because longitudinal in vivo recordings in behaving animals are technically demanding, and our primary goal was to follow and compare the dynamics across learning in WT and Shank3 KO mice. However, we recorded an additional habituation water session one day before CST1, which allows us to address (1) the decoding performance in naïve KO animals, and (2) whether higher correlated noise is already present in KO animals before learning. To this end, we first trained and tested the classifier on cross-registered data from the habituation (HAB) water session and the CST1 saccharin session across the CSonly (WT), CTA (WT), and CTA (KO) groups. Because this comparison is made across sessions, it is not exactly comparable to discrimination within a session, but it does show that GC responses in naïve KO animals can discriminate water from saccharin at baseline. These data are now included in Figure 6 – figure supplement 2 and are described in lines 329-338.

Interestingly, we also observed a significant reduction in the amplitude of suppressed responses during the habituation water session in KO compared to WT animals, as well as a trend toward higher stimulus-evoked neuronal coactivity (data included in Figure 2 – figure supplement 2). This suggests that reduced suppression is already present in GC of KO animals before learning, potentially contributing to slower CTA acquisition (now described in lines 209-215).

(2) The authors design an effective behavioral paradigm comparing consumption of water and saccharin and tracking extinction (Figure 3). This paradigm shows differences in licking across distinct behavioral conditions. For instance, during T1, licking to water strongly differs from licking to saccharin for both WT and KO. During T2, licking to water strongly differs from licking to saccharin only for WT (much less for KO), and licking to saccharin in WT differs from that in KO. These differences in taste sampling across conditions could contribute to some of the effects on neural activity and discriminability reported in Figures 5 and 6. That is, sucrose and water trials may be highly discriminable because in one case the mouse licks and in the other it does not (or licks much less). The author may want to address this issue.

This is an important point. As noted by the reviewer, active licking can modulate neuronal activity in the gustatory cortex independent of taste identity (Neese et al., 2022). Because our paradigm required animals to voluntarily sample tastants of different valences, motivated differences in licking are inherently tied to taste value, making it difficult to fully disentangle taste-evoked responses from lick-related activity.

However, taste exposure is known to induce prolonged neural responses that persist beyond the sampling phase (Juen et al., 2024). In our recordings, we included a 10second post-sampling epoch. We thus trained and tested classifiers using calcium traces during this post-sampling period; in particular, we divided the 10-second duration into five 2-second bins, matching the length of the tastant delivery phase, and analyzed the classifier built within each bin. We found that in both WT and KO animals, decoding performance was consistently above chance throughout the postdelivery period (now added to Figure 6 – figure supplement 1, lines 321-329), suggesting that decoding accuracies during sampling likely reflect taste rather than licking.

(3) Are there any omission trials following CTA? If so, they should be quantified and reported. How are the omission trials treated with regard to the analyses?

On the day following each CST session, animals underwent a water-only session (i.e., saccharin was omitted) to minimize context–malaise association. During these sessions, animals resumed licking both in the total lick counts and in the number of trials they engaged in, to levels comparable to pre-conditioning behavior. We did not observe significant differences between the WT and KO groups during these omission sessions. This point has been mentioned in the Methods section of the revised manuscript (lines 544-547).

(4) The authors describe the extinction paradigm as "alternative choice". In decision-making, alternative choice paradigms typically require 2 lateral spouts to report decisions following the sampling from a central spout. To avoid confusion, the authors may want to define their paradigm as alternative sampling.

We have revised this terminology to “alternative sampling” to avoid confusion with the classical alternative-choice paradigms.

(5) Figure 4 reports that CTA increases the proportion of neurons that consistently respond to saccharin and water across days. While the saccharin result could be an effect of aversive learning, it is less clear why the phenomenon would generalize to water as well. Can the authors provide an explanation?

Water and saccharin activated an overlapping population of neurons in GC. When we further quantified their tuning properties in the lifetime plots (Figure 4), we found that neurons responsive to both stimuli showed the most stable responsiveness across days, compared to neurons that responded only to saccharin or only to water (Author response image 1). Because the water-responsive and saccharin-responsive groups in Figure 4 both include this subset of dual-responsive neurons, this likely explains why both plots show increased reliability. This effect on reliability of single-cell responses is thus distinct from changes in the ability to discriminate between tastants at the population level (Fig. 6).

GC neurons responding to both water and saccharin are more stable during CTA extinction.

Lifetime plot showing significant responses of the same neurons responding to only water (blue), only saccharin (magenta), and to both saccharin and water (gold) across test sessions (T1-5) in the CTA (WT) group.

(6) The recordings are performed in the part of the anterior insular cortex that is typically defined as "gustatory cortex" (GC). Given the functional heterogeneity of the anterior insular cortex (AIC) and given that the authors do not sample all of the anteroposterior extent of AIC, I would suggest being more explicit about their positioning in GC. Also, some citations (e.g., Gogolla et al, 2014) refer to the posterior insular cortex, which is considered more inherently multimodal than GC. GC multimodality is typically associative in nature, as only a few neurons respond to sound and light in naïve animals.

Our stereotaxic coordinates targeted the conventional gustatory region within AIC (see revised manuscript Methods section, lines 489-490). We have revised the terminology throughout the manuscript to more explicitly reflect this anatomical positioning.

(7) It would be useful to add summary figures showing the extent of viral spread as well as GRIN lens placement.

Revised Manuscript Figure 1B shows a representative example of confirmed GRIN lens placement and the viral spread of GCaMP. In most cases, GCaMP expression is confined to GC, with minimal spread to the piriform cortex and along the injection track. Depth and GCaMP expression in GC were further validated during two-photon imaging.

(8) I encourage the authors to add Ns every time percentages are reported. How many neurons have been recorded in each condition? Can the authors provide the average number of neurons recorded per session and per animal?

We now included these numbers in the revised manuscript (lines 157-158, 162-163, 253, 268, 271-272).

(9) It looks like some animals learned more than others (Figure 1E or Figure 3C). Is it possible to compare neural activity across animals that showed different degrees of learning?

We thank the reviewer for this suggestion – we now show a significant correlation between the magnitude of CTA and the coactivity metric in Figure 1 Figure supplement 3; we elaborate in our Response to Reviewer #3 Public Review 1.

Reviewer #2 (Public review):

(1) Causality: The paper infers that increased correlated variability causes learning deficits, but no causal tests (e.g., optogenetic modulation of inhibition or interneuron rescue) are presented to confirm this.

Although we now provide data showing that correlated variability prior to learning is significantly correlated with the magnitude of CTA (see Response to Reviewer #1 Public Review 1above), we agree that we cannot infer causality without additional manipulations. While it might be possible to manipulate correlated variability by targeting inhibition within GC, optogenetic and chemogenetic manipulations of inhibition are likely to impact behavior through multiple mechanisms; for example, enhancing PV-interneuron activity in visual cortex profoundly impairs vision-dependent learning (Bissen et al. 2026, Leman et al. 2025). Thus, testing this would require finding a paradigm that specifically restores synchronization to WT levels without over-inhibiting the network, which is beyond the scope of the current study. We have rewritten the manuscript throughout to remove the inference of causality, and instead describe these two findings as being “associated” (see e.g. lines 95, 219-220, 379-382).

(2) Behavioural scope: The study focuses exclusively on taste aversion; generalisation to other flexible learning paradigms (e.g., reversal or probabilistic tasks) is not addressed.

Our study is focused on conditioned taste aversion (CTA) acquisition and extinction, which provides a well-established model for examining the formation and updating of aversive associative memories. We agree that cognitive flexibility encompasses a broad range of behavioral paradigms, and in the revised manuscript have sought to confined our conclusions to CTA. Whether the mechanisms identified here extend to other forms of flexible learning, such as reversal or probabilistic learning, or even to other sensory-stimulus-guided behavior, will require future investigation.

(3) Mechanistic insights: While providing interesting findings of altered sensory perception and extinction of learning-related signals in AIC, it offered nearly no mechanistic insights. This makes the interpretation, especially on how generalisable these findings are, difficult. Also, different reported findings are "potentially" connected, but the exact relation between increased correlated variability and faster loss of taste selectivity cannot be assessed.

In a new analysis we find that the coactivity level during CST1 showed a significant, positive correlation with the lick ratio (CST2/CST1); that is, the higher the initial coactivity during learning, the slower the CTA acquisition is (Figure 2 – figure supplement 3). This new piece of data (now added to revised manuscript lines 215218) provides a link between these two findings, and suggests that baseline coactivity levels in GC can influence the speed of CTA learning. We agree that this association does not imply causation, and have taken pains to avoid stating this.

Reviewer #3 (Public review):

(1) The authors don't make a causal link between the behaviour and AIC neurophysiology, both the percentage of suppressed cells and the coactivity measurements. For the % of suppressed cells, it seems that both WT and KO cells are suppressed in the transition between CST1 and CST2 (Figure 1L), yet only the WT mice exhibit CTA (at least by CST2). For the taste-elicited coactivity measure, it seems that there is an increase in coactivity from CST1 to CST2 in WT (Figure 2C - blue, although not statistically tested?), but persistently higher coactivity in KO. Is this change of coactivity in WT important for the expression of CTA? Plotting behavioral performance (from Figure 1G) against coactivity (from Figure 2C) for each animal would be informative.

This is a good suggestion (also made by the other reviewers), and we now show that the coactivity level during CST1 showed a significant, positive correlation with the lick ratio (CST2/CST1); that is, the higher the initial coactivity during learning, the slower the CTA acquisition is (Figure 2 – figure supplement 3).

(2) Shank3 KO cells already show an increase in baseline coactivity (Figure 2- figure supplement 1), and the authors never examine CS-only responses in the KO group, therefore making it difficult to determine whether elevated coactivity and noise correlations reflect a generalized AIC abnormality in Shank3 KOs (perhaps through impaired PV-mediated inhibition in insular cortex - Gogolla et al, 2014) that is not directly responsible/related to CTA?

We agree that the increased coactivity in KO animals likely reflects a general network defect in AIC before CTA learning. This baseline change is unrelated to the taste stimuli, because (1) the coactivity is already elevated before the animals receive the taste stimuli (that is, a baseline abnormality), (2) the neuronal responses to taste delivery during CST1 is indicative of CS-only responses, as it occurs before the injection of LiCl, when the associative learning process is initiated, and (3) a trend toward higher coactivity is already present in the habituation water session before CST1 (Figure 2 and Figure 2 – figure supplement 2).

We have clarified our description of these findings to avoid claiming that the increased coactivity “causes” poor learning performance (lines 379-382).

(3) How do the authors interpret the large range of lick ratios (Figure 1G) for WT (almost bi-modal distribution)? Is there a within-subject correlation with any of the neurophysiological measurements to suggest a relationship between AIC neurophysiology and behavioural expression of CTA?

See response to Point 1 above.

(4) Indeed, CTA appears to be successfully achieved for Shank3 KO mice delayed by 1 day, as the level of saccharin aversion during the first retrieval session (T1) is comparable between Shank3 KO and WTs. In this context, not extending the first part of the paradigm to include CST3 seems to be a missed opportunity. Doing so would have allowed for within-cell and within-subject comparison of taste-elicited pairwise correlation across the learning and to investigate the neural mechanism of delayed extinction in KOs more effectively.

We did not include a third CST session because when we analyzed the lick counts, KO animals already formed robust CTA after CST2 that was indistinguishable from WT animals. This suggests that the faster loss of CTA memory during extinction is due to a faster extinction process, rather than a weaker CTA memory from the outset. Adding a third CST could potentially lead to a memory that is harder to extinguish. Whether Shank3 KO mice would exhibit faster loss of memory in this scenario is an open question that would be interesting to explore in a future study.

(5) How to interpret Figure 5F: Absolute discriminability is lower for T5 for CTA WT and CTA KO compared to CS-only? Why would AIC neurons have less information on taste identity by the end of extinction than during the unconditioned (CS-only) condition? And if that is the case, how is decoding accuracy in Figure 6C higher in T5 for CTA WT vs CS-only?

We appreciate the reviewer's confusion about the discrepancy between our single-cell and population-level discriminability results. We speculate that in the CS-only state, individual AIC neuronal responses mostly reflect taste identity. However, after learning (in the CTA group), these neurons develop “mixed selectivity” (Tye et al., 2024), encoding not only identity but also the learned valence and extinction history. The lower single-cell discriminability after extinction (T5) in Figure 5F suggests that, although taste identity may remain constant, the learned history (e.g., "this taste used to be dangerous, but now it's safe") has shifted. This mixing of information makes each cell a weaker discriminator on its own.

However, the higher population decoding accuracy in Figure 6C demonstrates that the entire population of neurons can work together more effectively. The learning process could reorganize the neural ensemble in such ways that our support vector classifier (SVC) is able to identify and combine the relevant signals within the population, even when the valence of taste stimuli has changed, to better decode stimuli and outperform the non-learned state. This suggests that the brain shifted to a more robust, population-based coding strategy for complex, learned information, which is resistant to changes in selectivity at the single-cell level. The finding that population coding is robust to single-neuron variability has also been reported in other cortical regions (Montijin et al., 2016).

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

(1) Mechanistic experiments: Consider inhibitory neuron-specific imaging or manipulation (e.g., optogenetic enhancement of interneuron activity) to test whether restoring inhibition rescues learning flexibility.

We have addressed the limitation and potential issues for manipulating cortical inhibition in Response to Reviewer #2 Public review 1.

(2) Clarify limitations: Explicitly acknowledge the correlational nature of neuralbehavioural relationships in the Discussion.

We have removed language that implies a causal relationship throughout, and have emphasized the correlational nature of our findings in the Discussion section of our revised manuscript (lines 379-382)

(3) Enhance clarity: Simplify some dense methodological sections and expand figure legends to guide interdisciplinary readers.

We have adjusted the Methods section and figure legends as needed for better readability.

Individual Comments for Authors:

(1) L83-90: Confusingly written, not easy to understand for someone not knowing the paradigm in detail.

- What are the different stages? Memory encoding? Leaning? Extinction

- More reliable in taste responsiveness - what does that mean?

We have emphasized the behavior stages where each finding was observed in the revised manuscript (lines 82-91)

(2) L112: Not sure if these references support the "crucial", since they do not seem to be causal.

We have reworded this for accuracy (line 111).

(3) Figure 1: panels h and i in the heat maps, it looks like that in the KO animal, activity is more suppressed from CST1 to 2?

Panels j, l, m, and Figure 2: Neuronal suppression is already higher in CST1; therefore, there is no CTA effect but a general "perceptual" issue in the Shank3 model. The only effect seems to be a potential reduction in activation in CST2 in KO animals.

This point has been discussed in Reviewer #3 Public Review 2.

(4) Clarify in text. Especially with the sentence in the next paragraph, it might be confusing: "We wondered what other features of AIC activity during CTA acquisition might differ between WT and Shank3 KO mice."

We have rewritten this in the revised manuscript (lines 169-170).

(5) Clarify which are CTA-dependent and which are general (e.g., if writing suppression during CTA acquisition, it implies that it is related. But these changes were present before CTA.

We have clarified this in the revised manuscript (lines 209-215).

(6) Figure 4: Mainly shows a CTA-related increase in reliability in their taste responsiveness. This is not addressed anywhere else in the document and is not taken up in the discussion. How could it be related to the other findings, and what is its relevance? Please elaborate (e.g., in the discussion) or potentially remove?

We measured response reliability, as stabilization of stimulus-evoked responses has been reported in other sensory cortices across different learning tasks. Yet, it remained unclear whether CTA learning would induce similar changes in AIC. We took advantage of our longitudinal recording to address this question and believe that this piece of evidence will contribute to the research community that studies taste and learning in general. In addition, what is striking to us is that while the taste selectivity is degraded faster in KO animals, their response reliability is largely preserved. This suggests that these two sensory stimulus-related neuronal properties may involve distinct cellular and/or circuit mechanisms.

(7) Figure 5: Problematic to compare T5 between both groups, since T5 is lower than T4 in WT (against the trend) and T4 is an outlier in KO. e.g., if compared at T5, completely different results? Or why is there significance between T1 and T2 but not between T1 and T4 in KO? Could the authors address this point?

In Figure 5B, the slightly lower average for WT animals at T5 was driven by a single outlier, and there was no statistically significant difference between T4 and T5 (corrected post hoc t-test, WT, T4 vs. T5, p = 0.4097). Therefore, it does not contradict the trend toward an overall increase in nonselective neurons during CTA extinction. For KO animals, the lower average at T4 than T5 (corrected post hoc ttest, KO, T4 vs. T5, p = 0.0082) was intriguing, and one possible explanation is that neurons in the KO group might undergo more dynamic and variable changes in their responsiveness during CTA extinction, fluctuating before finally stabilizing.

Comparing T5 instead of T4 thus ensures that neuronal responsiveness is stabilized and reflects an “extinct” CTA memory more truly.

General Comments:

(1) While changes in SNR were observed in Shank3 models, the mechanism underlying decreased correlated variability has not been reported to date. Since decreased variability is usually associated with improved SNR ratio, it might be worth highlighting the distinction between "signal" and "noise" as separated in your analyses to make it more understandable for the reader.

We have described in the Results section what signal and noise correlations indicated and how they were separated in our analyses in both the Results and Methods section of the revised manuscript (lines 185-193, lines 673-681).

(2) What is the origin of the increased correlated variability?

We have discussed that reduced cortical feedback inhibition could be a potential source of increased correlated variability in the Discussion section of our revised manuscript (lines 370-375).

(3) Is the variability generally increased between trials (bigger fluctuations between trials for each neuron), or is the variability of each neuron similar, but they are just more correlated (more synced)?

Our pilot analysis did not detect any evident changes in the response variability for each neuron across trials; thus, we think that in KO animals, neuronal responsivity becomes more correlated and synchronized.

Reviewer #3 (Recommendations for the authors):

(1) Point in line 422-424: Rephrase the closing statement of the discussion as you have shown that mutant mice are actually able to update their behaviour (in fact faster) when the valence of the sensory input changes.

The “reduced ability to update behavior when the valence of a sensory input changes” refers to the finding that KO animals learned CTA more slowly; i.e., they were unable to timely adjust their behavior after malaise. We have rephrased this for clarity (line 448-449)

(2) The Figure 6 legend does not correspond to panels D and E in the figure. Νο I, J in figure.

We have fixed this mismatch in the revised manuscript.

Minor concerns:

(1) Cue/lick/taste-responding neurons greatly overlap and are not exclusively selective (Figure 1- figure supplement 2). Is there a genotype difference for the % of selective neurons (i.e., ones that only respond during cut/lick/taste) or the % of overlap?

When we quantified the stimulus responsivity in KO animals, we also identified neurons that were activated by cues, licks, or tastes. Their respective percentages and overlap did not differ significantly from those in the WT group, indicating that the modality of KO neurons across different sensorimotor cues is not compromised in the KO condition (Author response image 2).

Neurons in WT and Shank3 KO animals show comparable responsiveness to sensorimotor stimuli during conditioning.

(A) Percentage of neurons activated by the cue (left), lick movement (middle), and the tastant (right) in the CTA (KO) group (B) during the first conditioning session (CST1). (B) Venn diagram showing the overlaps among cue-, lick-, and tastant responsive neurons in (Figure 1 - figure supplement 2 C) and (A).

(2) For Figure 1: The authors could also express consumption as a % of consumed (trial-averaged licks) over the number of trials. It is mentioned that mice undergo daily training sessions consisting of 'approximately 30 trials' (line 114). This can give an indication of how strong the learning is between cta1 and cta2 and how strong the genotype difference is.

We are not sure if dividing trial-averaged licks over the number of trials would provide additional information, as the trial-average lick is already normalized to the number of trials.

(3) Figure 4: Why is there a different number of neurons in C vs G?

The figures B, C, D showed neurons that were activated by saccharin, and the figures F, G, H showed neurons that were activated by water. In all experimental groups, the numbers of neurons responsive to saccharin and water were different (i.e., B vs. F, C vs. G, D vs. H). The exact numbers were included in the corresponding figure legends in the revised manuscript.

(4) Figure 5B: The grey background box is moved to the left.

We kept the current figure format, as it effectively presents the mean, fitted mean, error bars, and individual animal data.

(5) In line 142: (1-2), (2-3), (3-4), the numbers in parentheses are confusing.

We have relabeled this as epoch 1-2, epoch 2-3, and epoch 3-4 in both text and figures for clarity (lines 145-146).

(6) Line 188: Do the authors mean noise correlations?

Rosenbaum et al. and Khoury et al. indeed measured correlated variability (noise correlation) in their study. On the other hand, Rothschild et al. did not specifically separate the noise from signal activities, which more likely reflect the coactivity measured in our case. We have rewritten this for accuracy (line 196).

(7) Where mentioning in the CS-only group, please explicitly state the CS-only WT group.

We have relabeled this throughout our revised manuscript.

(8) In lines 273-274: if the comparison is the reduction in discriminability being faster for the KO animals that had CTA, the correct comparison should be CSonly KO vs CTA KO.

We think that the better comparison to test how fast taste discriminability is reduced would be to perform post-hoc tests comparing T1 vs T2 within genotypes. We did not see significant changes between T1 and T2 in either genotype, which was reported in the figure legends of the reviewed preprint (lines 1140-1141).

Version 2

Abstract

The ability to update the valence of sensory perception to influence behavior is crucial for survival. A common phenotype in autism spectrum disorders (ASDs) is defects in sensory processing, but whether these defects impair flexible sensory encoding is largely unexplored. In particular, how genetic risk factors such as Shank3 deletion affect the adaptability of cortical taste processing and downstream behavior is unknown. To address this gap, we performed two-photon calcium imaging during a conditioned taste aversion (CTA) learning paradigm, an ethologically relevant form of associative learning that depends on taste processing in the gustatory cortex (GC), to examine how Shank3 knockout alters taste-related neuronal activity in GC and influences CTA learning. We found that GC neurons in Shank3 knockout mice exhibited reduced stimulus-evoked suppression and increased trial-to-trial correlated variability during the acquisition of CTA memory. These activity changes, which likely reduced signal-to- noise ratio in GC, were associated with slower CTA acquisition in knockout mice. In both genotypes, CTA learning enhanced, while subsequent extinction reduced, taste discriminability in GC, and both extinction and the associated reduction in discriminability were faster in knockout than in wild-type mice. Together, these results show that Shank3 loss is associated with destabilized cortical activity dynamics in GC, which may contribute to inefficient encoding and maintenance of learned taste aversion. These findings show that loss of Shank3 compromises the ability of animals to update behavior to incorporate negative outcomes, and suggest this loss of flexibility may be an important feature of monogenic ASDs.

Introduction

Ethological experiences can modify fundamental features of sensory coding, such as the valence of a taste, through changes in synaptic and network functions within sensory cortices (Harris and Mrsic-Flogel, 2013; Choi, 2018). This flexibility in sensory coding ensures that animals’ perceptions and downstream behaviors are updated to reflect current experiences and enhance survival. Sensory processing, including taste, is often defective in patients with autism spectrum disorders (ASDs) (Leekam et al., 2007; Avery et al., 2018; Scheerer et al., 2021; Goldschlager et al., 2025), and several rodent monogenic ASD models, including Shank3 knockout, exhibit sensory deficits resembling those observed in patients with ASD (Kloth et al., 2015; Copping et al., 2017; Monteiro and Feng, 2017; Delling and Boeckers, 2021). This altered sensory processing could contribute to downstream defects in flexible sensory coding in Shank3 knockout rodent models, but this possibility has not been directly tested. Here, we used conditioned taste aversion (CTA), an ethologically relevant form of associative learning that involves taste processing, to examine whether the learning and extinction of aversive taste associations are altered in Shank3 knockout mice.

CTA arises when a normally palatable taste is paired with agents (such as lithium chloride) that cause gastric malaise (Garcia et al., 1955); this aversive learning is fast (requiring one or two conditioning trials) and strongly suppresses the consumption of the conditioned tastant. Like other forms of aversive learning, CTA undergoes extinction when the tastant no longer predicts the adverse outcome (Berman and Dudai, 2001). Extinction occurs gradually over repeated exposures to the conditioned tastant alone; thus, during CTA and subsequent extinction, the valence of a tastant can shift from positive to negative and back to positive again in an ethologically relevant manner, providing a nice paradigm for studying the adaptability of sensory coding. CTA relies on the gustatory cortex (GC), a subregion within the anterior insular cortex (AIC), where much of taste coding occurs (Katz et al., 2002; Fletcher et al., 2017; Wang et al., 2018; Levitan et al., 2019; Vincis et al., 2020; Chen et al., 2021). The acquisition and extinction of CTA induce distinct forms of plasticity within GC (Escobar et al., 1998; Li et al., 2016; Rodríguez-Durán et al., 2017; Haley et al., 2020; Wu et al., 2021; Yiannakas et al., 2021; Kolatt Chandran et al., 2023), as internal representations of taste are reshaped (Moran and Katz, 2014; Lavi et al., 2018; Arieli et al., 2022; Raymond et al., 2024). Loss of Shank3 affects multisensory integration in the insular cortex (Gogolla et al., 2014) and impairs several cellular forms of cortical synaptic and intrinsic plasticity (Bozdagi et al., 2010; Wang et al., 2011; Kouser et al., 2013; Raynaud et al., 2013; Tatavarty et al., 2020; Wen and Turrigiano, 2021), raising the possibility that loss of Shank3 might impair CTA learning and extinction.

In this study, we employed two-photon calcium imaging and a head-restrained CTA paradigm to compare GC activity in Shank3 knockout and wild-type mice during learning and extinction. At baseline and during memory encoding, GC neurons in Shank3 knockout mice were less suppressed by taste stimuli, with increased correlated variability that likely reduced the signal-to-noise ratio critical for taste processing within GC. This was associated with slower formation of CTA, and although they eventually formed strong CTA, this aversive memory was extinguished more rapidly than in wild-type mice. Following the CTA acquisition phase, GC neurons in both groups became more reliable in their taste responsiveness; however, this reliability was accompanied by a gradual reduction in single-cell taste selectivity and a convergence of population activity as CTA extinction progressed. Like CTA extinction, this loss of taste selectivity and convergence of population activity was faster and more pronounced in Shank3 knockout than in wild-type mice. Finally, the ability to discriminate between tastants using population activity was enhanced by CTA and then maintained during extinction in wild- type mice, but was rapidly lost during extinction in the Shank3 knockout condition. This work demonstrates that loss of Shank3 degrades the discriminability of single-cell and population activity in GC, which may contribute to slower learning and faster extinction of CTA. More broadly, this work shows that flexible sensory coding is altered in Shank3 knockout mice, and provides insight into how maladaptive sensory coding impacts associative learning in ASD.

Results

Shank3 knockout diminishes taste-elicited GC neuronal suppression and perturbs CTA acquisition

Loss of Shank3 is known to affect some aspects of sensory processing (Chen et al., 2020; Tatavarty et al., 2020; Balasco et al., 2022; Ortiz-Cruz et al., 2022), but whether it disrupts ethologically relevant associative learning paradigms such as CTA that rely on sensory processing is unknown. To examine this, we performed longitudinal two-photon calcium imaging of GC neurons from wild-type (WT) and Shank3-knockout mice (KO, Δexon 13-16, Peça et al., 2011) in a head-restrained CTA learning paradigm (Figure 1A-C). Using a gradient-index (GRIN) lens (Miura et al., 2020; Deng et al., 2021; Staszko et al., 2022) and a virally expressed calcium indicator (jGCaMP8s) driven by the pan-neuronal promoter Synapsin (Zhang et al., 2023), we were able to visualize calcium signals of individual neurons within deep layers of the dysgranular and agranular GC, subregions that contribute to the expression of CTA (Figure 1B, C, Yiannakas et al., 2021; Kolatt Chandran et al., 2023).

Shank3 knockout diminishes taste-elicited GC neuronal suppression and perturbs CTA acquisition.

(A) The head-restrained aversive taste learning paradigm. CS: conditioned stimulus, 25 mM saccharin, US: unconditioned stimulus, lithium chloride (LiCl, 0.15M, 2% body weight). HAB: habituation session, CST: conditioning session, T: retrieval test session. (B) Example image showing the GRIN lens placement on top of the gustatory cortex (GC) (green: GCaMP8s, magenta: NeuN). Scale bar: 500 μm. (C) Example image showing the calcium signals recorded from individual GC neurons. Scale bar: 50 μm. (D, E, F) Trial- averaged licks during the first (CST1) and second (CST2) conditioning sessions in the CS-only (WT) group (D), the CTA (WT) group (E), and the CTA (KO) group (F) (CS-only (WT): paired t-test, p = 0.1201; CTA (WT): paired t-test, p = 0.0094; CTA (KO): paired t-test, p = 0.0490). Here and below where behavior data are presented: CS-only (WT), n = 7 animals; CTA (WT), n = 9 animals; CTA (KO), n = 9 animals. (G) The trial-averaged lick ratio of CST2 to CST1 in CTA (WT) and CTA (KO) groups. (WT vs. KO: unpaired two-tailed t-test, p = 0.0123). (H, I) Heatmaps showing z-scored changes in calcium responses (ΔF/F) from neurons that are responsive to tastant delivery during conditioning in CTA (WT) (H, CST1: 189 cells, CST2: 172 cells) and CTA (KO) (I, CST1: 66 cells, CST2: 52 cells) groups. 1: cue onset; 2: spout in place; 3: tastant delivery starts; 4: tastant delivery ends. The 3-second taste delivery is marked by yellow dashed lines. In each heatmap, cells are sorted in descending temporal order based on whether their activity is activated or suppressed. Here and below where calcium imaging data are presented: CS-only (WT), n = 7 animals; CTA (WT), n = 8 animals; CTA (KO), n = 7 animals. (J) Calcium traces averaged across suppressed neurons in the CTA (WT) and CTA (KO) groups during CST1 (left) and CST2 (right). (K) Calcium traces averaged across activated neurons in the CTA (WT) and CTA (KO) groups during CST1 (left) and CST2 (right). (L) Change in the percentage of suppressed neurons from CST1 to CST2 (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.0218, session: p = 0.0393, interaction: 0.7819). (M) Change in the calcium fluorescence (ΔF/F) of suppressed neurons from CST1 to CST2. Dashed lines mark the value 0 (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.0011, session: p = 0.7356, interaction: 0.8672). (N) Change in the percentage of activated neurons from CST1 to CST2 (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.1391, session: p = 0.2590, interaction: 0.9705). (O) Change in the calcium fluorescence (ΔF/F) of activated neurons from CST1 to CST2 (WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.3404, session: p = 0.2664, interaction: 0.1295). Error bars in this figure and below indicate s.e. (standard error of the mean).

We then trained water-restricted, head-restrained mice (WT or KO littermates) to lick a motorized spout in daily sessions, each consisting of approximately 30 trials (Chen et al., 2021). This trial-based structure enabled us to control which tastant animals received in each trial, as well as the duration and timing of its delivery, while simultaneously recording the corresponding neuronal calcium signals in GC (Figure 1A, C). Prior to conditioning, animals underwent two imaged habituation sessions (HAB), wherein only water was delivered (see methods), allowing us to record baseline water-evoked GC neuronal activities before CTA learning. To ensure that animals learned CTA effectively, we then subjected habituated animals to two conditioning sessions (CST1 and CST2); within each session animals received, on average, 30 deliveries of the novel tastant saccharin (conditioned stimulus, CS), followed by a LiCl injection (unconditioned stimulus, US) (referred to as the “CTA group”). A separate group of wild-type animals received saline instead of LiCl and served as the “non-learned” comparison (referred to as the “CS-only (WT) group”). In subsequent retrieval test sessions that spanned five days (T1-5), we measured the expression and extinction of CTA by giving animals saccharin (the conditioned tastant) and water (a neutral tastant) in an alternative- sampling task (Chen et al., 2021; Kogan and Fontanini, 2024). While animals in the CS-only (WT) group maintained their preference for saccharin across the first and second conditioning sessions (CST1 and CST2, Figure 1D), WT mice that received LiCl significantly reduced licks to saccharin between CST1 and CST2 (Figure 1E). KO animals also showed a reduction in their saccharin consumption between CST1 and CST2 (Figure 1F), but this reduction was significantly less than for WT mice (Figure 1G).

To ensure reduced CTA during CST2 in KO mice was not an artifact of measuring aversion without a choice between saccharin and water, we used a modified paradigm where a separate group of animals received an alternative-sampling test after CST1 (Figure 1 – figure supplement 1A). In this paradigm as well, KO mice showed a significantly higher preference for saccharin across concentrations compared to WT littermates (Figure 1 – figure supplement 1B); further, comparable lick counts in WT and KO mice suggest no major differences in the thirst state or motivation between groups (Figure 1 – figure supplement 1C, D).

Since taste responses are processed in GC (Katz et al., 2002), we wished to know whether saccharin-elicited activity in GC during CTA acquisition is different in WT and KO animals. We identified neurons that were activated or suppressed during various phases of CST1 and 2, and (as observed previously, Chen et al., 2021) found that GC neurons could be influenced by acoustic cues (epoch 1-2), lick movements (epoch 2-3), and tastant exposure (epoch 3-4, Figure 1 – figure supplement 2A-C). Most of these neurons responded to more than one sensorimotor stimulus (Figure 1 – figure supplement 2D), highlighting the multi-modal and integrative nature of the insular cortex (Samuelsen et al., 2012; Gogolla et al., 2014; Vincis and Fontanini, 2016; Gogolla, 2017). In most GC neurons, the strongest trial-averaged responses were tied to tastant delivery (Figure 1H, I, Figure 1 – figure supplement 2A, B; between yellow dashed lines marked 3 and 4); in this study, we focus on these taste-elicited responses. In both WT and KO mice that underwent CTA, we detected neuronal populations that were either suppressed (Figure 1H, I, J) or activated (Figure 1H, I, K) by tastant presentations during CST1 and CST2. There were no significant changes between CST1 and 2 for either the CTA or the CS alone groups (Figure 1 – figure supplement 2E-J). Strikingly, however, there was less suppression in response to saccharin in the KO mice; a smaller fraction of neurons were suppressed (Figure 1L; percentage averaged from 8 (WT) and 7 (KO) animals: WT-CST1: 40%, WT-CST2: 31%, KO-CST1: 22%, KO-CST2: 14%), and the average magnitude of suppression was smaller (Figure 1J, M; magnitude averaged from 8 (WT) and 7 (KO) animals: WT-CST1: -1.96, WT-CST2: -1.91, KO-CST1: -0.98, KO-CST2: -0.81). The population size and magnitude of activation of activated neurons were also smaller, but this effect did not reach statistical significance (Figure 1N, O; percentage averaged from 8 (WT) and 7 (KO) animals: WT-CST1: 26%, WT-CST2: 23%, KO-CST1: 18%, KO-CST2: 15%. Averaged magnitude: WT-CST1: 3.82, WT-CST2: 3.92, KO-CST1: 3.84, KO-CST2: 2.28). These data reveal that Shank3 KO mice have less cortical suppression during tastant delivery, and exhibit reduced or slowed acquisition of CTA.

Shank3 knockout increases the correlated variability in GC during CTA acquisition

We wondered if, in addition to reduced suppression in response to saccharin, other features of GC activity might differ between WT and Shank3 KO mice. The correlated variability between neurons – the trial-to-trial co-fluctuation of their activity over repeated presentations of the same stimulus – has been shown to limit the information-encoding capacity of neuronal networks (Moreno-Bote et al., 2014; Bartolo et al., 2020; Rumyantsev et al., 2020; Kafashan et al., 2021; Hazon et al., 2022). To determine whether correlated variability differs between WT and KO mice, we first characterized the trial-to-trial coactivity between GC neurons by measuring the pairwise correlations of their mean calcium responses across CS (saccharin) presentations (see methods). There was a rightward shift in the cumulative distribution function of these pairwise correlations in KO compared to WT animals, which was apparent in both conditioning sessions (Figure 2A, B). Comparing the mean pairwise correlation calculated by animal confirmed a persistent increase in the average strength of coactivity in the KO mice across conditioning sessions (Figure 2C).

Shank3 knockout increases the correlated variability in GC during CTA acquisition.

(A, B) Cumulative distribution of taste-elicited coactivity (Pearson coefficients) between neurons during CST1 (A) and CST2 (B). Dashed lines indicate the shuffled data (CST1: WT, n = 6201 pairs, KO, n = 2645 pairs, WT vs. KO, Kolmogorov-Smirnov test, p < 0.0001; CST2: WT, n = 8476 pairs, KO, n = 2847 pairs, WT vs. KO, Kolmogorov-Smirnov test, p < 0.0001). (C) Change in the average taste-elicited coactivity from CST1 to CST2, plotted by animal. The dashed line indicates the chance-level correlation. (WT vs. KO, repeated- measures mixed ANOVA, genotype: p = 0.0142, session: p = 0.4198, interaction: 0.4146). (D) Schematic showing how the signal and noise correlations are computed. (E, F) Change in the signal correlation from CST1 to CST2, plotted by neuronal pair (E) or by animal (F) (E: WT: CST1, n = 6201 pairs, CST2, n = 8476 pairs; KO: CST1, n = 2645 pairs, CST2, n = 2847 pairs, two-way ANOVA, genotype: p = 0.1134, session: p = 0.0009, interaction: 0.5469; F: WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.8143, session: p = 0.4007, interaction: 0.8213). (G, H) Change in the correlated variability from CST1 to CST2, plotted by neuronal pair (G) or by animal (H) (G: two-way ANOVA, genotype: p < 0.0001, session: p = 0.2274, interaction: 0.6112; H: WT vs. KO, repeated-measures mixed ANOVA, genotype: p = 0.0617, session: p = 0.4017, interaction: 0.7188). (I) Average taste-elicited coactivity in CST1, grouped by inter-neuronal distance and plotted by animal (WT vs. KO, repeated-measures mixed ANOVA, genotype, p = 0.0050, distance, p = 0.7643, interaction: p = 0.5895).

Neuronal coactivity could result from “signal” correlations, which reflect the similarity of average stimulus-tuning properties between neurons, and “noise” correlations that measure trial-to-trial activity co-fluctuations between neurons independent of stimulus tuning (Panzeri et al., 2022). To determine if the “noise” component is the driving force of the increased coactivity observed in the KO mice, we performed a different correlation analysis that allowed us to separate signal-induced correlations from the residual (noise) correlations (Marks and Goard, 2021). For a given neuron, we first obtained the stimulus-induced signal by averaging response traces (3 seconds) across trials (Figure 2D, top). We then estimated the neuron’s trial-to-trial variability (‘noise’) by subtracting this averaged ‘signal’ from individual trials and concatenating the residuals (Figure 2D, bottom). The signal correlation was then quantified by calculating the pairwise correlations of the averaged signal, while the noise correlation was calculated as the pairwise correlations in the trial-to-trial variability. While the signal correlation was reduced from CST1 to CST2, there was no difference between WT and KO mice (Figure 2E, F). By contrast, the correlated variability was significantly higher in the KO mice (Figure 2G, H). Both coactivity and correlated variability between neurons generally decrease when the intraneuronal distances grow wider (Rothschild et al., 2010; Rosenbaum et al., 2017; Khoury et al., 2022). To determine the spatial extent of enhanced correlations in Shank3 KO mice, we measured correlated variability as a function of the inter-somatic distance between neuron pairs (Figure 2I); strikingly, the increased coactivity was observed even in neuronal pairs > 300 μm apart (Figure 2I), indicating a network-wide, rather than local, origin. Thus, in the Shank3 KO mice, there is a widespread increase in correlated variability between GC neurons, suggesting there is more “noise” in gustatory cortical circuits.

Next, to determine if this coactivity is stimulus-dependent, we repeated the correlation analysis using the mean calcium responses two seconds prior to cue onset (Lee et al., 2020). Measuring the trial-to-trial correlations of this spontaneous activity showed that coactivity was enhanced in KO animals even at baseline (Figure 2 – figure supplement 1A-C). Furthermore, as for taste-elicited activity, this increased baseline coactivity had a broad spatial extent (Figure 2 – figure supplement 1D). To determine whether these differences are present before learning, we next performed the same analyses on recordings during the habituation session one day before CST1, using water as the tastant (Figure 1A, see methods). We again observed a significant reduction in the amplitude of suppressed response during the habituation session in the KO compared to the WT group (Figure 2 – figure supplement 2D), as well as a trend toward higher stimulus-evoked neuronal coactivity (Figure 2 – figure supplement 2E). Lastly, we examined the relationship between the coactivity metric and the magnitude of CTA. We found that the degree of coactivity at the onset of learning (CST1) was significantly correlated with less CTA (expressed as the CST2/CST1 lick ratio) (Figure 2 – figure supplement 3). Taken together, these analyses demonstrate that Shank3 KO mice have increased network-wide, stimulus-independent correlated variability within GC, and suggest that this correlated variability may contribute to impaired acquisition of CTA.

Extinction of CTA in Shank3 knockout mice is faster than in wild-type littermates

The reduced expression of CTA in KO mice after the first conditioning session (Figure 1) could be due to a reduction or a delay in the learning process. To differentiate between these possibilities, we measured taste aversion after the second round of conditioning (after CST2), by comparing the ratio of licks to saccharin and water during the first retrieval test session (T1, Figure 1A). We found that both WT and KO mice exhibited strong and comparable CTA during T1 (Figure 3A-D), indicating that CTA formation is slower but still significant in KO mice.

Extinction of CTA in Shank3 knockout mice is faster than in wild-type littermates.

(A, B, C) Trial-averaged licks of saccharin (violet) and water (blue) in the CS-only (WT) (A), CTA (WT) (B), and CTA (KO) (C) groups throughout the test sessions (CS-only (WT): repeated-measures ANOVA, taste, p = 0.2530, session, p = 0.4158, interaction, p = 0.4515; CTA (WT): repeated-measures ANOVA, taste, p < 0.0001, session, p < 0.0001, interaction, p < 0.0001; CTA (KO): repeated-measures ANOVA, taste, p = 0.0002, session, p < 0.0001, interaction, p < 0.0001). (D) Trial-averaged lick ratios of saccharin to water throughout the test sessions in CTA (WT) and CTA (KO) groups (WT vs. KO: repeated-measures mixed ANOVA, genotype, p = 0.0069, session, p < 0.0001, interaction, p = 0.0691, post-hoc Tukey test, WT vs. KO, T1 p = 0.8669, T2 p = 0.0032, T3 p = 0.2705, T4 p = 0.9014, T5 p = 0.9531). (E, F) A fitted logistic curve (top) and its derivative (bottom) describing the relationship between lick ratios and test sessions in an example WT mouse (E) and an example KO mouse (F). (G) Average rate of change (derivative) at each test session. The triangle indicates the data mean. (WT vs. KO: repeated-measures mixed ANOVA, genotype, p = 0.0271, session, p < 0.0001, interaction, p = 0.0003; post-hoc Tukey test, WT vs. KO, T1 p < 0.0001, T2 p = 1.0000, T3 p = 1.000, T4 p = 1.000, T5 p = 1.000).

As expected, WT animals underwent gradual memory extinction between T1 and T5 (Berman and Dudai, 2001; Yiannakas and Rosenblum, 2017), with their preference for saccharin slowly returning close to baseline by T4-5 (Figure 3B, D). In contrast, KO mice underwent much faster extinction, with their saccharin consumption returning to baseline by T2 (Figure 3C, D). To calculate the rate of extinction, we fitted a logistic curve to the lick-ratio data for each animal and estimated the rate of memory extinction by taking the derivative of the fit (Figure 3E, F, Kogan and Fontanini, 2024). This analysis revealed distinct extinction dynamics between WT and KO mice (Figure 3E, F): as expected, the lick ratio of saccharin to water increased gradually over time in WT (Figure 3E, top) while rising steeply in KO mice, with a much higher peak rate (Figure 3F, G). Thus, Shank3 KO mice not only form CTA more slowly but also undergo faster CTA extinction than their WT littermates.

CTA learning increases the response reliability of individual GC neurons

The stimulus responsiveness of individual sensory neurons is prone to drift over time (Deitch et al., 2021; Marks and Goard, 2021; Schoonover et al., 2021). We wondered whether CTA might stabilize taste-elicited cortical responses, as has been suggested in other sensory modalities (Poort et al., 2015; Glas et al., 2021; Taylor et al., 2021; Biane et al., 2023; Pancholi et al., 2023). To test this, we identified GC neurons that could be recorded in each retrieval test session (Figure 4A, see methods) and tracked their daily responsiveness to taste. We first focused on cells that were activated by saccharin in at least one session and found that in the CS-only (WT) group, a majority of neurons dropped in and out of the population responses (Figure 4B). In contrast, both WT and KO CTA groups had a higher proportion of neurons reliably activated by saccharin across days (Figure 4C, D). While no neurons in the CS-only (WT) group were active across all five sessions, 16 (WT) and 15 (KO) percent of neurons in the CTA conditions were responsive throughout (WT: 11/70 neurons, KO: 6/39 neurons, Figure 4E). The same pattern was found for GC neurons that were responsive to water, an independent and neutral taste modality that induces distinct neuronal activities in GC (Figure 4F-I, Katz et al., 2001; Rosen et al., 2010). Hence, associative learning enhances the reliability of GC neurons’ responsiveness in both genotypes.

CTA learning increases the response reliability of individual GC neurons.

(A) Example images showing neurons present in all five retrieval test sessions (labeled green). Scale bar: 20 μm. (B, C, D) Lifetime plot showing significant responses of the same neurons across test sessions (T1-5) in the CS-only (WT) group (B, n = 42 cells), the CTA (WT) group (C, n = 70 cells), and the CTA (KO) group (D, n = 39 cells). Violet labels activated states, and white labels nonactivated states. (E) Percentage of neurons active for 1, 2, 3, 4, and 5 sessions in (B-D) (Chi-square test, p < 0.0001). (F, G, H) Lifetime plot showing the water- responsiveness of the same neurons across test sessions (T1-5) in the CS-only (WT) group (F, n = 43 cells), the CTA (WT) group (G, n = 63 cells), and the CTA (KO) group (H, n = 36 cells). Blue labels activated states, and white labels nonactivated states. (I) Percentage of neurons active for 1, 2, 3, 4, and 5 sessions in (F-H) (Chi-square test: p < 0.0001).

The stimulus selectivity of individual GC neurons degrades faster in Shank3 knockout mice during memory extinction

The feature selectivity of individual neurons within primary sensory cortices is dynamic and experience-dependent (Chéreau et al., 2020). We reasoned that taste selectivity – the “breadth of tuning” (Levitan et al., 2019) – of individual GC neurons might evolve during CTA extinction, and that this process might be accelerated in Shank3 KO mice to promote faster extinction. To test this, we identified neurons activated by water or saccharin delivery, and in test sessions estimated the percentage of neurons that responded to one or both stimuli (Figure 5A). Our analysis revealed a comparable proportion of nonselective GC neurons (∼40 percent; percentage averaged from 8 (WT) and 7 (KO) animals: WT, 41%, KO, 36%) in both genotypes during the first test session (T1, Figure 5B). During extinction, the percentage of nonselective neurons gradually increased in the WT group and reached ∼50 percent by T5 (average: 49%; Figure 5B); this increase was faster and stronger in KO mice, reaching ∼65 percent by T5 (average: 65%; Figure 5B).

The stimulus selectivity of individual GC neurons degrades faster in Shank3 knockout mice during CTA extinction.

(A) Calcium traces of example neurons activated by saccharin only (left), water only (middle), or both (right). Violet marks the calcium responses to saccharin, and blue marks the responses to water. (B) The percentages of nonselective neurons across test sessions and genotypes. Each data point represents one animal. The colored bar marks the data-averaged value and its standard error. The square indicates the fitted mean in each condition based on the linear mixed-effects model, with session and genotype set as fixed parameters and individual animal as random effect (WT vs. KO, linear mixed-effects model, genotype (in reference to WT): KO, β = -5.146, p = 0.554, session (in reference to T1): T2, β = 5.410, p = 0.438, T3, β = 11.692, p = 0.093, T4, β = 14.433, p = 0.038, T5, β = 8.045, p = 0.248, interaction: KO x T2, β = 16.729, p = 0.103, KO x T3, β = 13.466, p = 0.189, KO x T4, β = -5.420, p = 0.597, KO x T5, β = 21.409, p = 0.037, random effect, variance, 74.187; corrected post hoc t-test, KO, T1 vs. T2, p = 0.0326, KO, T4 vs. T5, p = 0.0082, WT, T4 vs. T5, p = 0.4097). (C, D, E) Histograms comparing the distributions of the discriminability index between T1 and T2 (top) and between T1 and T5 (bottom) in the CS-only (WT) group (C, T1: 243 cells, T2: 238 cells, T5: 250 cells, F-test with Bonferroni correction, T1 vs. T2 p = 0.5979, T1 vs. T5 p = 0.2110), the CTA (WT) group (D, T1: 355 cells, T2: 382 cells, T5: 358 cells, F-test with Bonferroni correction, T1 vs. T2 p = 0.6687, T1 vs. T5 p = 0.0001), and the CTA (KO) group (E, T1: 212 cells, T2: 212 cells, T5: 241 cells, F-test with Bonferroni correction, T1 vs. T2 p < 0.0001, T1 vs. T5 p < 0.0001). (F) Comparison of absolute discriminability index across CS-only (WT), CTA (WT), and CTA (KO) groups. Each data point indicates a single-cell value, grouped by animal. Black bars indicate the “predicted” subject- level mean value from the mixed-effects model accounting for both fixed (session, genotype) and random (animal) effects. The colored bars mark the “predicted” marginal means from fixed effects (session and genotype) (linear mixed-effects model, group (in reference to CS-only (WT), T1): CTA (WT), β = 0.1058, p = 0.1538, CTA (KO), β = 0.2442, p = 0.0018; session (in reference to CS-only (WT), T1): T2, β = 0.0271, p = 0.4598, T5, β = -0.0035, p = 0.9239; interaction: CTA (WT) x T2, β = -0.0577, p = 0.2203, CTA (WT) x T5, β = -0.1082, p = 0.0213, CTA (KO) x T2, β = -0.2393, p < 0.0001, CTA (KO) x T5, β = -0.2856, p < 0.0001). Planned post hoc comparisons are tested using linear contrasts on the fitted model with Benjamini- Hochberg correction for multiple comparisons (T2 vs.T1: CTA (WT), p = 0.6455, CTA (KO), p = 0.9742; WT vs. KO: T1, p = 0.6455, T2, p = 0.0301, T3, p = 0.1040).

While binarizing calcium signals into “active” and “non-active” is widely used to describe single-cell stimulus-elicited responses (Zhang and Li, 2018; Gehrlach et al., 2019; Wang et al., 2020; Chen et al., 2021; Staszko et al., 2022; Biane et al., 2023), it ignores changes in response amplitude that could encode stimulus-related information. We therefore performed a complementary analysis and calculated a discriminability index based on the amplitude differences between each neuron’s trial-averaged calcium responses to saccharin and water (see methods, Banerjee et al., 2020; Lee et al., 2021; Kong et al., 2023). A more positive index indicates the neuron has a higher response preference for saccharin, while neurons with a more negative index respond more strongly to water. We first quantified the spread of the distribution by pooling the discriminability indices from all recorded neurons, where a narrowing of the distribution indicates that neurons within the population become more similar in their response preference. In the CS-only (WT) group, the variance was similar for the first and last test sessions (Figure 5C). In WT mice that underwent CTA, there was a gradual reduction in the variance (narrowing) of the distribution from T1 to T5 (Figure 5D), and this decrease was already apparent during T2 in KO mice (Figure 5E). Calculating the absolute discriminability index (which reflects the degree of a neuron’s selectivity regardless of the taste stimulus it responds to) revealed that by T5, the mean absolute index was significantly smaller in the CTA group compared to the CS-only (WT) group (CS-only (WT) vs. CTA (WT) at T5, p = 0.0213, Figure 5F). This reduction in taste discriminability was again faster and more pronounced in KO mice (CTA (WT) vs. CTA (KO) at T2: p = 0.0301, Figure 5F). This reduction in discriminability was not due to changes in the number and response amplitude of both taste-activated and suppressed GC neurons, as these remained stable throughout extinction (Figure 5 – figure supplement 1). Thus, although the responses of individual GC neurons become more reliable after CTA, they gradually provide less information about taste identity during CTA extinction, and this process is accelerated in the KO condition.

Learning improves population coding of taste in GC, but this enhancement rapidly decays in Shank3 knockout mice

To determine whether the experience-dependent changes in taste discriminability are reflected in population-level activity, we calculated the Euclidean distance between population vectors in response to water or saccharin exposures; this is a widely used measure of the similarity between population neuronal responses to a pair of stimuli (Levitan et al., 2019; Arieli et al., 2022). In the CS-only (WT) group, the population responses to water and saccharin remained highly similar, as shown by the consistently small distance between the two (Figure 6A). By comparison, CTA roughly doubled the distance (dissimilarity) between the two taste- elicited population responses, and while this distance decreased during extinction, it remained higher than for the CS-only (WT) group (Figure 6A). To determine whether these population differences are sufficient to reliably classify tastants, we trained a support vector classifier to predict the taste class (water or saccharin) based on the ΔF values in the neuronal feature space (see methods, Bishop, 2006; Woods et al., 2020; Raymond et al., 2024). We trained and evaluated a classifier for each animal within each session to capture potential behavioral variability (Twarkowski et al., 2022). While decoding accuracy was close to chance in the CS- only (WT) group, CTA learning led to significantly better classification performances across sessions (Figure 6B-C). The distance between population responses to water and saccharin was reduced faster and to a larger degree in KO mice (Figure 6D, linear mixed-effects model: T1 vs. T2 in WT, β = -0.099, p = 0.697; T2 WT vs. KO, β = -0.901, p = 0.015; post hoc paired t- test: T1 vs. T2 in KO, p = 0.009), and this could also be seen in degraded decoding accuracy in these mutant mice (Figure 6E, F).

Learning improves population coding of taste in GC, but this enhancement rapidly decays in Shank3 knockout mice.

(A) Comparison of the Euclidean distance between population responses to water and saccharin between CTA (WT) and CS-only (WT) groups throughout test sessions (linear mixed-effects model, group (in reference to CS-only (WT)): CTA, β = 0.992, p = 0.005, session (in reference to T1): T2, β = -0.065, p = 0.760, T3, β = -0.046, p = 0.829, T4, β = 0.101, p = 0.636, T5, β = 0.013, p = 0.950, interaction: CTA x T2, β = -0.033, p = 0.910, CTA x T3, β = -0.282, p = 0.336, CTA x T4, β = -0.680, p = 0.020, CTA x T5, β = -0.441, p = 0.132, random effect, variance, 0.312). (B) Confusion matrices showing the prediction performances in T1 (left) and T5 (right) in an example CTA (WT) mouse. (C) Comparison of decoding accuracies of the linear classifiers trained and tested in each session for the CS-only (WT) and the CTA (WT) groups. The dashed line indicates the chance-level predictions (CS-only (WT) vs. CTA (WT): two-way ANOVA, group, p = 0.0001, session, p = 0.6637, interaction, p = 0.3273). (D) Comparison of the Euclidean distance between population responses to water and saccharin between CTA (WT) and CTA (KO) groups throughout test sessions (linear mixed-effects model, genotype (in reference to WT): KO β = 0.225, p = 0.625, session (in reference to T1): T2, β = -0.099, p = 0.697, T3, β = -0.328, p = 0.195, T4, β = - 0.578, p = 0.022, T5, β = -0.427, p = 0.091, interaction: KO x T2, β = -0.901, p = 0.015, KO x T3, β = -0.621, p = 0.094, KO x T4, β = -0.674, p = 0.069, KO x T5, β = -0.618, p = 0.095, random effect, variance, 0.538; post hoc t-test, KO, T1 vs. T2, p = 0.0089). (E) Confusion matrices showing the prediction performances in T1 (left) and T5 (right) in an example CTA (KO). (F) Comparison of decoding accuracies of the linear classifiers trained and tested in each session for the CTA (WT) and the CTA (KO) groups. (WT vs. KO: two-way ANOVA, genotype, p = 0.0078, session, p = 0.7293, interaction, 0.7757). (G) Heatmaps of decoding accuracies using the classifiers trained on T1 responses and tested across sessions. Each row indicates an animal. (H) Quantification of the decoding accuracies in (I), comparing CTA (WT) and CTA (KO) groups (linear mixed-effects model, genotype (in reference to WT): KO, β = - 1.417, p = 0.846, session (in reference to T1): T2, β = 0.031, p = 0.995, T3, β = -6.286, p = 0.214, T4, β = -7.545, p = 0.136, T5, β = -10.031, p = 0.047, interaction: KO x T2, β = -17.732, p = 0.017, KO x T3, β = -14.260, p = 0.054, KO x T4, β = -8.013, p = 0.279, KO x T5, β = -8.261, p = 0.264, random effect, variance, 96.479; post hoc t-test, KO, T1 vs. T2, p = 0.0063).

To rule out that these decoding changes were driven by changes in licking patterns (Neese et al., 2022), we leveraged the previous report that taste exposure induces prolonged neuronal responses that persist beyond the licking (sampling) phase (Juen et al., 2024), and trained classifiers using calcium traces during the 10-second post-sampling period in each trial. We divided the 10-second duration into five 2-second bins, matching the length of the tastant delivery phase, and analyzed the classifier built within each bin. We found that in both WT and KO animals, decoding performance was consistently above chance throughout the post-delivery period (Figure 6 – figure supplement 1), suggesting that decoding accuracies during sampling likely reflect taste discriminability rather than licking patterns. We also wondered if the ability to discriminate saccharine from water was already impaired in KO animals prior to learning. To gain some insight into this question we trained classifiers on neuronal activities recorded during the habituation (water) session one day before CST1 and during the CST1 (saccharin) session (see methods). Interestingly, classifiers trained with data on separate days produced decoding accuracy above chance level across genotypes and conditions, with the KO group showing slightly poorer performance (Figure 6 – figure supplement 2). While this analysis is not directly comparable to discriminability within a session (as in Figure 6), it does suggest that GC responses in naïve KO animals can discriminate water from saccharin at baseline.

Lastly, we examined the stability of population-level taste representations in GC during memory extinction (across sessions) by using the classifiers trained on T1 responses to predict taste classes in the subsequent sessions (T2-5). In the CS-only (WT) group, the classifier performed poorly across sessions (Figure 6G). In contrast, in the CTA (WT) condition, the cross-session classifier performed well during the early stages of memory extinction, and the accuracy gradually decreased (T1 vs. T5, p = 0.037, Figure 6G, H). Finally, consistent with our previous findings, in the CTA (KO) condition, the decoding accuracy declined sharply from T1 to T2 (Figure 6G, H). In summary, population-level taste representations in GC become more discriminative after learning. During CTA extinction, these representations progressively diverge from those during early memory retrieval, and in KO animals, this divergence is faster.

Discussion

Flexible sensory coding ensures that perceptions are updated to reflect their value to an organism, to enable adaptive behaviors. While sensory coding deficits have been reported in mouse models of ASD, including Shank3 KO (Monday et al., 2023, Wang and Feldman, 2024), it is unknown whether flexible coding is impaired in ethological learning paradigms such as CTA. Here, we used 2-photon calcium imaging and a head-restrained CTA learning paradigm to address this question in Shank3 WT and KO mice. We show that neuronal ensembles in GC, the cortical region critical for taste processing, have higher correlated noise and lower taste discriminability in Shank3 KO compared to WT littermates, and these changes are correlated with slower CTA learning. Thus, Shank3 KO animals require repeated exposure to aversive experiences to enable updating of taste valence. Once CTA is established, responses of GC neurons in both genotypes are able to discriminate between different tastants at both the single-cell and population levels. However, during CTA extinction, taste discriminability in Shank3 KO mice degrades faster than in their wild-type counterparts, and this coincides with faster extinction. This slower formation and faster extinction of CTA indicates that loss of Shank3 compromises the ability of animals to update taste valence and downstream behaviors to reflect negative outcomes.

One possible contributor to slower CTA acquisition is the increased correlated variability between GC neurons in Shank3 KO mice, which is likely to reduce signal-to-noise ratio during sensory encoding (Moreno-Bote et al., 2014; Bartolo et al., 2020; Rumyantsev et al., 2020; Kafashan et al., 2021; Hazon et al., 2022). Reduced cortical feedback inhibition has been shown to increase correlated variability in the cortex (Stringer et al., 2016), and we detected reduced neuronal suppression in GC of Shank3 KO mice, consistent with known impairment of cortical interneuron function after loss of Shank3 (Lee et al., 2017; Chen et al., 2020; Pagano et al., 2023), raising the possibility that the increase in correlated variability stems from weakened inhibition within GC. Similar heightened noise correlations have been observed in sensory cortices in other rodent ASD models, including mice with cell-type-specific deletion of Syngap1 and MeCP2 (Banerjee et al., 2016; Zhao and Kwon, 2023). A striking feature of the correlated noise we observed in GC of Shank3 KO mice was the large (>300 µm) spatial scale, suggesting it could reflect correlated activity of divergent inputs to GC. Regardless of the source, we found a significant correlation between the magnitude of the correlated variability and reduced CTA formation across animals, suggesting that this noise contributes to slower CTA encoding, although establishing causality will require additional manipulations. Notably, a recent study showed slower CTA acquisition in a pharmacologically-induced rat model of autism (Pal et al., 2025), suggesting that impairment of CTA also generalizes across ASD models. Taken together with previous findings, our observations suggest that a reduced signal- to-noise ratio driven by high correlated variability in the sensory cortex is a common mechanism contributing to maladaptive sensory processing in ASD.

GC neurons exhibit a wide range of tuning properties (Fletcher et al., 2017; Levitan et al., 2019; Chen et al., 2021; Staszko et al., 2022), and we found that, similar to neurons in the primary visual and somatosensory cortices (Poort et al., 2015; Khan et al., 2018; Chéreau et al., 2020), the stimulus selectivity (response preference) in GC can be dynamically adjusted at the cellular level, from being selective to nonselective during CTA extinction. While CTA learning can still initiate selective response preferences in GC neurons of Shank3 KO mice, this stimulus selectivity was rapidly degraded in concert with the faster extinction of CTA memory. The maintenance and extinction of a faithful CTA memory rely on both Hebbian and homeostatic synaptic plasticity mechanisms within GC (Escobar et al., 1998; Li et al., 2016; Rodríguez-Durán et al., 2017; Haley et al., 2020; Wu et al., 2021). Because Shank3 KO is known to perturb several of these critical plasticity processes (Bozdagi et al., 2010; Raynaud et al., 2013; Tatavarty et al., 2020), rapidly reduced stimulus selectivity of GC neurons could result from defects in one or more of these local plasticity mechanisms. An alternative, though not mutually exclusive, possibility is that the rapid loss of stimulus selectivity and CTA memory results from changes in other brain regions that project to GC, such as BLA and the ventral tegmental area (VTA) (Levitan et al., 2020; Gil-Lievana et al., 2022). Hebbian plasticity can be induced at synaptic inputs from BLA to GC during CTA learning (Escobar et al., 1998; Rodríguez-Durán et al., 2017; Haley et al., 2020), and dopaminergic projections from VTA facilitate the induction of NMDAR-dependent LTP within GC (Rodríguez-Durán et al., 2023). Thus, Shank3 KO could disrupt function in these upstream brain regions to impair plastic changes in GC during extinction. This possibility is consistent with the finding that Shank3 down-regulation in VTA accelerates extinction of socially induced conditioned place preference (CPP) memory (Bariselli et al., 2018).

During CTA extinction, population responses to the neutral and conditioned taste stimuli in GC gradually become more similar. Because CTA learning can be suppressed by the familiarity of a tastant (De la Casa and Lubow, 1995), we were not able to compare the population responses to different taste stimuli before and after CTA learning; however, even after extinction these population responses were more distinct in the CTA than the CS-only (WT) group, suggesting that taste coding in GC does not revert to its naïve state even when the behavioral aversion is gone. This is consistent with in vivo electrophysiology demonstrating that, during CTA extinction, the population-level taste coding in GC continuously evolves but does not return to the pre-CTA state (Moran and Katz, 2014). In contrast, a recent study (Raymond et al., 2024) reported that the population responses to conditioned and appetitive taste stimuli converge to a much greater degree than we observed after CTA extinction. A possible explanation for this difference is that, because we used two conditioning sessions with a novel tastant, while Raymond et al. used a single conditioning session with a familiar tastant, our paradigm likely elicits stronger CTA, which may then initiate a more enduring change in discriminability in WT animals. Interestingly, in Shank3 KO animals, faster CTA extinction is tied to faster and stronger convergence of population taste-evoked responses in GC, suggesting that learning-induced changes in sensory encoding are more easily reversed.

Our finding that taste representations remain discriminative in GC even after extinction in WT animals raises an important question: what taste information does this GC population activity encode? In the hippocampus, representations of odor “identity” remain stable independent of learning, while those encoding odor “values” are susceptible to extinction (Biane et al., 2023). In GC, ensemble activity encodes both the identity and palatability of taste stimuli, as well as related decision-making, in sequential temporal epochs (Katz et al., 2001; Sadacca et al., 2016; Livneh et al., 2017; Lavi et al., 2018; Chen et al., 2021; Staszko et al., 2022; Kogan and Fontanini, 2024; Raymond et al., 2024). The population GC activity we analyzed (3-second duration after delivery) is likely to carry information about both taste identity and taste palatability, or “valence.” If learning-induced changes largely reflect changes in palatability (as expected, Moran and Katz, 2014), this would explain the enhanced decoding performances of classifiers after CTA, when water and saccharin move from similar to dissimilar palatability (Staszko et al., 2022).

Flexible processing and assignment of value to incoming sensory information in the cortex enable animals to update their perceptions and modify behaviors that are critical for survival, and disruptions in this process could contribute to maladaptive behaviors in neurodevelopmental disorders. Here, we show that in Shank3 KO mice there is a lower signal- to-noise ratio in GC, and this is associated with slower CTA learning and faster extinction. This rapid extinction is associated with rapid loss of taste discriminability within GC, and this response instability ultimately compromises the ability to decode taste identity from GC responses. More broadly, these data suggest that altered sensory processing in some forms of ASD may impair the ability to update behavior when the valence of a sensory input changes.

Materials and methods

Resource Availability

Lead contact and materials availability

Further information and requests for resources and reagents should be directed to the lead contact, Gina G. Turrigiano (turrigiano@brandeis.edu).

Experimental Model and Subject Details

All experimental procedures were approved by Brandeis University Institutional Animal Care and Use Committee and followed the National Institute of Health guidelines for the Care and Use of Laboratory Animals. C57BL/6 and Shank3B knockout (Δexon 13-16, JAX: 017688) mouse strains were obtained from the Jackson Laboratory, and the colonies were maintained in Foster Biomedical Research Labs at Brandeis University. Adult mice (8-16 weeks old) were used in experiments performed in this study. All animals were grouped with their littermates or housed individually once they entered the behavioral training paradigm. In all cases, animals were housed in a humidity- and temperature-controlled environment and entrained to a 12- hour light-dark cycle (light phase from 7:00-7:00), with ad libitum access to food and water unless described otherwise. Mice of each sex were assigned equally to each experimental condition. In this paper, 7, 9, and 9 animals were assigned to the CS-only (WT), CTA (WT), and CTA (KO) groups in the two-conditioning behavior experiments (Figures 1 and 3). For calcium imaging analysis, 1 and 2 animals were excluded from the CTA (WT) and CTA (KO) groups, respectively, due to a lack of sufficient GCaMP expression (Figures 1–6). In the separate single-conditioning paradigm, 13 and 13 animals were used in the CTA (WT) and CTA (KO) groups.

Viral Vectors

The AAV serotype 1 expressing pGP-AAV-syn-jGCaMP8s (Addgene: 162374) was obtained from Addgene.

Stereotaxic Surgery

Mice (8-10 weeks old) were anesthetized with a constant airflow containing 2% isoflurane throughout the procedure. The skull was exposed, and a craniotomy was created above the right GC. AAV expressing GCaMP8s (titer: ∼1.2 x 1013, 1:3 dilution, 800 nL in total) was slowly infused into GC (anterior-posterior (AP) with reference to bregma: 1.0 mm, medial- lateral (ML): - 3.8 mm, dorsal-ventral (DV): with reference to the brain surface: -2.0 mm) through a micropipette connected to a microinjector (Narishige, MO-10), at a rate of approximately 200 nL/min. To allow the virus particles to spread, the pipette was kept in place for another 5 minutes before being slowly withdrawn from the site. Two weeks after the virus injection, the mice (10-12 weeks old) then underwent the implantation of a gradient-index (GRIN) lens and a customized titanium head bar. Mice were anesthetized as described above, and a dose of dexamethasone (2 mg/kg) was administered subcutaneously at the beginning of the surgery to minimize brain swelling. Once the skull was revealed, a craniotomy slightly larger than 0.5 mm x 0.5 mm was made at the site of the previous virus injection. To create a leading track, a 25-gauge blunt needle was slowly lowered until it reached above GC (AP: + 1.0 mm, ML: -3.8 mm, DV: -1.8 mm), remained in place for five minutes and was withdrawn from the site. A GRIN lens (diameter: 0.5 mm, length: ∼4 mm, CLHS050GFT126, GoFoton) was then gradually lowered through a customized holder until its bottom surface rested on top of GC (AP: + 1.0 mm, ML: -3.8 mm, DV: -2,0 mm). The lens was secured to the skull by Vetabond and the UV-curable adhesive, after which a titanium head-bar was installed. Metabond tinged with black ink was applied to further cement the lens and the head bar while covering and sealing the skull. After the operation, animals were kept and monitored individually in their home cages, and a daily dose of meloxicam (1-2 mg/kg) was given subcutaneously for at least three days to reduce brain inflammation. Animals were allowed for at least one week of recovery before starting behavioral training.

Head-restrained CTA Learning Paradigm

One day before the mice (13-15 weeks old) began training in the head-fixed position, ad libitum water access was removed from their home cages to motivate their engagement in the subsequent tasks. The restricted water access was then given daily through the motorized spout during the training task or in the home cage for thirty minutes (described in detail below). Each training day, animals were placed on a customized wheel (Livneh et al., 2017) with their heads restrained and underwent a thirty-minute training session consisting of 30-35 trials. In each trial, a two-second monotonal acoustic cue (2 kHz) was initiated to signal the onset of the trial, with a motorized spout simultaneously moving to the location close to the animal’s mouth. Mice then learned to perform five dry licks in order to receive the liquid through the spout. The gravity-based liquid delivery, once initiated, lasted 3 seconds, after which the spout was moved away from the animal. A 10-second intertrial rest followed before the next trial started. A trial was considered failed if the animal did not reach the threshold to initiate the liquid delivery within sixty seconds after the spout was in place.

Habituation

During the habituation stage, water was used as the tastant. If animals refused to lick the spout throughout the session or their body weight dropped over 15 percent of the initial weight, an additional ad libitum access to water was granted for 30 minutes in the home cage. On average, it took 10 days for the animals to become familiarized with the task, reaching consistent performance without failed trials within thirty minutes. By the end of the habituation, the spout became the only daily source of liquid the animals received, and their body weight stabilized. Starting from the last two habituation sessions, GC neuronal activities were recorded by two-photon microscopy (see below) while animals were performing the task. The recording during these habituation sessions ensured that animals were accustomed to the mechanical operation of image acquisition, while allowing us to obtain “baseline” water-evoked GC activity data prior to learning.

Conditioning

Once habituated, animals then received two conditioning sessions to induce the formation of conditioned taste aversion (CTA). On each conditioning day, animals underwent the training session with the same scheme as in habituation, except that saccharin (25 mM, CS) was given instead of water. Five minutes after the session, a dose of lithium chloride (LiCl, 0.15 mM, 2% body weight, US) was intraperitoneally injected to induce gastric malaise. Afterward, animals remained in place for another 25 minutes before returning to the home cage. To overcome the potential association between malaise and the context, which might confound animals’ subsequent performance, animals were allowed to receive water in a recovery session (in the same environment) the next day after each conditioning; during this time, animals usually resumed their engagement in the task.

CTA retrieval and extinction testing

After the second recovery session, we subjected animals to consecutive test sessions for five days to measure the retrieval and extinction of CTA memory. In each test session, water was given for the first three trials to engage animals in the task. Afterward, water and saccharin were delivered in a semi-randomized order such that animals could not predict the identity of the tastant in the upcoming trial while overall receiving similar numbers of trials that delivered water and saccharin.

Single-conditioning CTA paradigm

A modified taste learning paradigm was conducted to examine whether Shank3 knockout mice expressed taste aversion when presented with choices after one conditioning. After the first conditioning session and the subsequent recovery session, animals were immediately subjected to a retrieval test session in which water and saccharin at various concentrations (25 mM, 10 mM, 5 mM, 1 mM) were delivered in a semi-randomized order. Trial-averaged licks and lick ratios were quantified as described below.

Behavior Data Analysis

Trial-averaged lick calculation

Throughout the paradigm, licks were detected by a capacitive sensor (MPR121, Adafruit) attached to the spout, which generated TTL signals recorded with the customized code and time-stamped to the frames during two-photon image acquisition. To obtain trial-averaged licks of a tastant (Figure 1 and 3), total lick counts during the 3-second delivery were averaged across trials delivering that tastant in a given session. In the cases where the licks of saccharin and water were compared (Figure 3), the ratio was further calculated by dividing the trial-averaged licks of saccharin over that of water.

Logistic curve fitting

A logistic curve was fitted to the trial-averaged lick ratios as a function of time (session) to describe the temporal dynamics of the licking behavior (Kogan and Fontanini, 2024). The logistic function was fitted using the curve_fit function in the Python scipy model and is defined as: (1) f(t)=L1+e−k(t−t0) where f(t) is the fitted lick ratio at time t; L is the maximum lick ratio; k denotes the growth rate, and t0 indicates the inflection point. The derivative of the logistic function was calculated as: (2) f′(t)=Lke−k(t−t0)(1+e−k(t−t0))2

Immunohistochemistry

Animals were deeply anesthetized with isoflurane and perfused with 4% paraformaldehyde (PFA). The whole brain was extracted, post-fixed in 4% PFA for 24 hours, and sliced on a vibrating microtome (Leica Vibratome VT 1200s). Coronal brain slices (50 µm) containing GC were collected; to verify the placement of the GRIN lens, the slices with the visible track left by the lens were selected, washed three times with PBS, permeabilized with the blocking buffer (5% goat serum/3% BSA/0.3% Triton X-100 in PBS) at room temperature for 2 hours, and incubated with primary antibodies diluted in the blocking buffer at 4°C overnight. To confirm the expression of GCaMP8s, chicken anti-GFP (1:1000, Aves Labs) and mouse anti-NeuN (1:500, MAB-377, Millipore) were used. On the next day, the slices were washed three times with PBS and incubated with secondary antibodies diluted in the buffer containing 5% goat serum and 3% BSA at room temperature for 3 hours (goat anti-chicken Alexa-488 and goat anti-mouse Alexa-594, 1:400, Thermo-Fisher). After another three washes with PBS, the slices were mounted on the slides using the DAPI-Fluoromount-G mounting medium (SouthernBiotech), cover-slipped, and stored at 4°C until imaging.

Confocal Imaging

To verify the expression of GCaMP8s and the placement of the GRIN lens, stained slices were imaged on a laser scanning confocal microscope (Zeiss LSM880) using the tiling scan under a 20x objective, with a frame size of 1024 pixels x 1024 pixels. For each hemisphere, the images were stitched online using the ZEN Black software (Zeiss) and then analyzed using ImageJ/FIJI (NIH, US).

Two-photon Imaging

Acquisition

Two-photon imaging was conducted using a resonant-scanning two- photon microscope (Neurolabware, 31 frames per second; 512 pixels x 512 pixels) with a 10X objective (Thorlabs, NA: 0.50, air, working distance: 7.77 mm, zoom factor: 2.8). GCaMP8s was excited at 920 nm using a tunable ultrafast laser (SpectraPhysics Insight X3, 50-80 mW measured at the front aperture of the objective). Fields of view of all imaging were selected approximately 200 µm below the bottom surface of the GRIN lens. On average, 750 frames (25 seconds) were obtained for each trial.

Processing and registration

For each imaging session, frames from all successful trials were included and concatenated into a series. In each movie, the movement was corrected, the regions of interest (ROI) corresponding to individual neurons in the field of view were identified, and the raw fluorescence from the cell bodies and neuropils was extracted using Suite2P (Pachitariu et al., 2017). The processed images were then inspected, and if needed, the ROIs were manually selected or excluded using the Suite2P GUI so that only the cell bodies were used for further analysis. Only the ROIs near the center of the field of view were included to avoid any potential peripheral distortion of fluorescence by spherical aberration (Sattin et al., 2024). The time series of neuropil-corrected calcium signals of individual neurons were obtained by subtracting the time series of neuropil fluorescence (scaled by 0.7) from that of the raw fluorescence of cell bodies. In the case where multi-day cell registration was required (i.e., identification of the same cells across sessions, Figure 4 and 6), Spatial footprints of each neuron were extracted from the field of view of each session with in-house customized code and were aligned across sessions using the probabilistic modeling method implemented in CellReg (Sheintuch et al., 2017). Centroid distance was recommended by CellReg in most cases, and a probability threshold of 0.5 was used.

Cell inclusion criteria

Cells were identified by the classifier built and trained in Suite2p. Once identified, each cell’s extracted somatic and neuropil calcium traces were then inspected; if the contamination from neuropils was severe (i.e., the magnitude of neuropil calcium signals was consistently higher than that of somatic signals), that cell was excluded from the analysis.

Calcium Imaging Data Analysis

Calculation of ΔF/F0 and stimulus responsiveness

Z-scored changes in calcium signals were used to represent the temporal activity dynamics of individual neurons in all analyses. In each trial, neuropil-corrected calcium traces of the same duration (total 656 frames) were first extracted, which consisted of 155 frames of baseline (5 seconds before the cue onset), 62 frames of the acoustic cue (2 seconds), 15 frames of dry licks (∼ 0.5 seconds), 112 frames of tastant delivery (3.5 seconds), and 312 frames of post-delivery rest (10 seconds). On average, it took animals no more than 0.5 seconds (15 frames) to reach the threshold of five dry licks; however, during the early stages of memory extinction, animals occasionally hesitated or even failed to participate in some trials, prolonging the dry-lick period to varying degrees. In such cases, we only included 15 frames starting from the initiation of the first dry lick to ensure that the lengths of the time series were consistent across animals, trials, and sessions. For each neuron in a given trial, we first computed its changes in fluorescence (ΔF/F0) by subtracting the mean fluorescence of baseline (F0) from the fluorescence at each time point and dividing the result values by F0. Next, we calculated the z-scores using the formula (F(t) – Fbaseline) / SD, where F(t) represents the ΔF/F0 at time t, and Fbaseline and SD denote the mean and standard deviation of the ΔF/F0 values during the baseline. To determine whether a neuron is responsive to a taste stimulus in a given session (Figure 1, 2, 4, 5), we compared the mean z-scored activity during the taste delivery (last 2 seconds) to the mean z- scored baseline activity (2 seconds before the cue onset) across trials using paired t-test. A neuron was categorized as “activated” and “suppressed” during the tastant delivery if there was a significant increase and decrease in activity compared to the baseline, respectively. We omitted the first second of the tastant delivery when calculating the mean activity to avoid a potential carryover from the dry lick responses (Chen et al., 2021). In the analyses that used the amplitude of the trial-averaged response (z-scored ΔF/F0, Figure 1, 2, 4, 5, 6), we first averaged the z-scored calcium traces across trials and then calculated the values of the mean activity during tastant deliveries as described above.

Coactivity analysis

The calculation of taste-elicited coactivity patterns between neuronal pairs was adapted from Twarkowski et al., 2022. Within each session, trials of the same tastant deliveries were selected, and the Pearson correlation coefficient between two neurons’ mean z-scored ΔF/F0 values during the tastant delivery across these trials was calculated. To determine the stimulus-independent baseline coactivity, the mean z-scored ΔF/F0 values during the 2-second period before the cue onset were used to compute the Pearson correlation coefficient.

Pairwise signal and noise correlations

The signal and noise correlations were calculated as previously described (Marks and Goard, 2021). The signal correlation was obtained in a given session by calculating the Pearson correlation between two neurons’ full trial-averaged activity traces during the tastant delivery. To estimate the correlated variability (noise), we first subtracted a neuron’s trial-averaged activity trace from its z-scored activity trace in each trial. The subtracted traces (residuals) from all trials within a session were then concatenated into a whole series representing trial-to-trial variability in the neuron’s activities. The Pearson correlation coefficient was calculated between two neurons’ concatenated, subtracted traces to represent correlated variability.

Single-cell discriminability index

The discriminability index was adapted from the previously reported method (Kong et al., 2023) and indicated a neuron’s response preference between taste stimuli (water or saccharin). In a given session, the discriminability index for each neuron is defined as: (3) Discriminability index=μ1−μ212(σ12+σ22) where µ1 and µ2 denote the trial-averaged activity amplitude averaged across saccharin and water trials, respectively, and the denominator is the pooled standard deviation. A positive index indicates that the neuron responds more preferentially to saccharin, while a more negative index indicates a higher response preference for water.

Euclidean distance

We calculated the pairwise Euclidean distance to estimate the similarity between population responses to saccharin and water (Levitan et al., 2019; Chen et al., 2021; Raymond et al., 2024). In a given test session, for each animal, we constructed the trial-averaged ΔF values of all neurons into two population vectors, one responding to saccharin and the other to water. The Euclidean distance between these two vectors is then defined as: (4) Normalized Euclidean distance=∑i=1n(v1i−v2i)2n Where v1 and v2 denote the population vector in response to saccharin and water, respectively, and n is the number of neurons included in the vectors. To mitigate the influence on Euclidean distance by the number of neurons included in the calculation (i.e., the dimension), we normalized the distance by the square root of the dimension (Levitan et al., 2019).

Decoding analysis

We trained a support vector classifier for each animal to predict the tastant class (water or saccharin) in a given session, using the SVC function with a linear kernel in the Python scikit-learn module (Woods et al., 2020; Twarkowski et al., 2022; Raymond et al., 2024). We first constructed population response vectors for each trial (labeled as saccharin or water), and these vectors constituted the dataset for model training and testing. The linear decoder is defined as: (5) L(t)=θ(wTv(t)+b) where L(t) is the predicted tastant class, and v(t) is the population response vector for the trial t. θ(z) is the Heaviside step function that ensures hard classification, while w denotes the vector that contains weights assigned to each neuron, and b is the bias term (Bishop, 2006). We then used 10-fold cross-validation, a standard method for model validation and evaluation, to train and test the data. Cross-validation produced an averaged accuracy score that indicated the classifier’s performance. To calculate chance-level prediction, we shuffled the labels and trained the classifier as described above for 1000 iterations. This shuffling process resulted in an average 50-percent accuracy score by chance. To test the stability of GC taste representations, we used the classifier trained on the data in test 1 to predict the tastant classes in the subsequent sessions and calculated the accuracy score accordingly.

Decoding analysis prior to CTA learning (baseline)

To evaluate how well GC population activity decoded taste stimuli before CTA learning, support vector classifiers were trained and tested on taste-evoked activities of GC neurons registered in both the last habituation session (water) and CST1 session (saccharin). We used the dataset combined across two sessions for this purpose because, to prevent the effects of saccharin pre-exposure on CTA learning, we did not subject the animals to the alternative-sampling tasks (i.e., two taste stimuli within a session) until the test sessions (i.e., after learning).

Quantification and Statistical Analysis

All statistical analyses were performed using the Python scipy, statsmodels, or pingouin packages. To assess the main effects and interactions in experiments comparing the same animals or neurons across sessions under different conditions, a paired t-test or repeated- measures mixed ANOVA was used unless stated otherwise. For experiments requiring specific comparisons between conditions, a linear mixed-effect model was fitted to the data, and the relevant coefficient was extracted. If the comparison could not be directly inferred from the coefficient, a post hoc t-test was performed. The significance levels were labeled by asterisks (*): *: P < 0.05; **: P < 0.01; ***: P < 0.001; ****: P < 0.0001.